---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
library(Rducks)
library(future)
```

# Rducks

[![R-CMD-check](https://github.com/sounkou-bioinfo/Rducks/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rducks/actions/workflows/R-CMD-check.yaml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Rducks)](https://sounkou-bioinfo.r-universe.dev/Rducks)

Rducks registers R functions as DuckDB SQL functions using a DuckDB C
extension, including a small set of unstable DuckDB C extension API functions.
The extension records the DuckDB database instance handle at initialization and
keeps an extension-owned connection associated with that runtime. UDF inputs and
outputs move through explicit execution plans:

- `arrow_r`: reference path using DuckDB Arrow C Data plus nanoarrow/R.
- `arrow_c`: native extension path for supported scalar and vectorized calls
  with direct DuckDB-vector materialization. The vectorized `RCV` evaluator uses
  direct native chunk materialization/writeback, not the old Arrow/R helper
  bridge.
- `arrow_ipc`: process-isolated R execution using Arrow IPC request/result
  payloads through generic Future or experimental persistent mirai workers.

The user-facing UDF semantics are separate from the execution plan:
`mode = "scalar"` means one R call per logical row; `mode = "vectorized"` means
one R call per DuckDB chunk.

## Quick start

```{r getting-started}
library(DBI)
library(duckdb)
library(Rducks)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(con, threads = "single")

reg_hello <- rducks_register(
  con,
  name = "r_hello",
  fun = function() "hello from R",
  args = NULL,
  returns = VARCHAR,
  mode = "scalar"
)

dbGetQuery(con, "SELECT r_hello() AS message")

reg_plus_one <- rducks_register(
  con,
  name = "r_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "scalar"
)

dbGetQuery(con, "SELECT r_plus_one(41.0) AS x")
```

Use `args = NULL` for a zero-argument scalar UDF. `null_handling` only affects
SQL NULL inputs, so the default is appropriate for `r_hello()`.

`rducks_register()` returns an `rducks_registration` object that records the
connection, normalized signature, and registration options. You do not need to
keep this object for the UDF to keep working in DuckDB. Registering the same SQL
name/signature again replaces the callable implementation in the shared DuckDB
database catalog.

### Release and disconnect

`rducks_release(con)` detaches connection-local Rducks state such as the current
default plan and R-side runtime anchor. It is non-destructive: it does not drop
DuckDB catalog functions and does not release closures still owned by native
catalog metadata. Call it before `DBI::dbDisconnect(con)` when you want
deterministic R-side cleanup. Rducks does not provide `rducks_unregister()`;
registered catalog UDFs and their preserved R closures are intentionally retained
for the database/runtime lifetime or until the same SQL name/signature is
replaced.

```{r release-example}
release_con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(release_con, threads = "single")
rducks_release(release_con)
dbDisconnect(release_con, shutdown = TRUE)
```

### Arrow conversion setting

`rducks_enable()` sets DuckDB's `arrow_lossless_conversion=true` setting on the
user connection, and the Rducks extension also sets it on its own internal
DuckDB connection. Rducks uses DuckDB's Arrow C Data conversion APIs for several
execution plans, and this setting tells DuckDB to preserve DuckDB-specific type
identity in Arrow metadata when possible instead of exporting only a more generic
Arrow representation. This matters for types such as `UUID`, `BIT`,
`HUGEINT`/`UHUGEINT`, `TIME WITH TIME ZONE`, and JSON aliases. It does not
change stored DuckDB values or SQL evaluation; it only affects DuckDB-to-Arrow
conversion used at the Rducks boundary.

## Scalar and vectorized modes

Scalar mode calls the R function once per DuckDB row. Vectorized mode calls the
R function once per DuckDB chunk with one R vector/list-column per declared
argument.

```{r vectorized-mode}
reg_vec_plus_one <- rducks_register(
  con,
  name = "r_vec_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "vectorized",
  side_effects = TRUE
)

dbGetQuery(con, "SELECT sum(r_vec_plus_one(i::DOUBLE)) AS x FROM range(5) AS t(i)")
```

In vectorized mode, `null_handling = "default"` evaluates only rows with no
top-level SQL NULL inputs and scatters SQL NULLs back into the result.
`null_handling = "special"` passes all rows using the same NA/NULL shapes as
scalar mode. The return length must match the number of evaluated rows.
Vectorized mode currently requires at least one declared argument.

A tiny benchmark with `bench` can show the call-shape difference for simple R
work. The result is illustrative rather than a performance guarantee.

```{r vectorized-benchmark}
bench::mark(
  scalar = dbGetQuery(con, "SELECT sum(r_plus_one(i::DOUBLE)) AS x FROM range(10000) AS t(i)"),
  vectorized = dbGetQuery(con, "SELECT sum(r_vec_plus_one(i::DOUBLE)) AS x FROM range(10000) AS t(i)"),
  iterations = 3,
  check = FALSE
)[, c("expression", "median", "itr/sec", "mem_alloc")]
```

## Execution plans

Execution plans choose the default marshalling implementation and concurrency
contract for future registrations through a connection. Registration remains
semantic: name, function, mode, types, NULL handling, exception handling, and
side-effect flag. The plan active at `rducks_register()` freezes the UDF's
native evaluator/marshalling metadata; later plan changes do not retarget
already-registered UDFs.

| Plan | Scalar | Vectorized | Notes |
| --- | --- | --- | --- |
| `arrow_r + serial` | implemented | implemented | reference implementation |
| `arrow_r + inproc_concurrent` | implemented | implemented | queued same-process callbacks; R API work stays on the recorded main R thread |
| `arrow_c + serial` | implemented | implemented | direct native evaluator tokens `RC`/`RCV` |
| `arrow_c + inproc_concurrent` | implemented | implemented | queued same-process callbacks with direct `arrow_c` marshalling |
| `arrow_ipc + multiprocess_parallel` | implemented | implemented | Arrow IPC request/result payloads through `ipc_future_pool` or experimental `ipc_mirai_pool`; evaluator token `RIPC` |

`arrow_r + serial` is the semantic reference. Other implemented plans are tested
against it and do not silently fall back to another marshalling path. Use
`rducks_explain_udf()` to inspect the plan and native counters for a registered
UDF.

```{r explain-udf}
rducks_explain_udf(con, "r_vec_plus_one")[, c(
  "name", "mode", "plan_id", "native_marshalling",
  "evaluator", "arrow_r_chunks", "arrow_c_chunks", "arrow_ipc_chunks"
)]
```

## In-process queued execution

`rducks_enable_inproc()` lets DuckDB worker threads enqueue same-process R calls
while the recorded main R thread drains the queue. This is for safety and
liveness, not automatic speed: R API work still runs on the main thread.

```{r inproc-queue}
rducks_register(
  con,
  name = "r_inproc_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE
)

rducks_enable_inproc(con, threads = 4, external_threads = 1)
dbExecute(con, "CREATE TEMP TABLE inproc_input AS SELECT i::DOUBLE AS x FROM range(200000) AS t(i)")
dbGetQuery(con, "SELECT sum(r_inproc_plus_one(x)) AS total FROM inproc_input")
rducks_inproc_stats(con)[, c("submitted", "executed", "pending_max", "main_drain_batches")]
rducks_disable_inproc(con, threads = 1)
```

## Multiprocess Arrow IPC execution

`arrow_ipc + multiprocess_parallel` uses Arrow IPC bytes for chunk tasks and
results. Scalar mode loops over rows inside the worker process; vectorized mode
calls the R function once per chunk inside the worker process. The default
provider is generic Future; `ipc_provider = "mirai"` uses persistent mirai
workers.

```{r arrow-ipc-plan}
future::plan(future::multisession, workers = 2)

plan <- rducks_execution_plan(
  "arrow_ipc", "multiprocess_parallel",
  future_packages = "stats",
  future_timeout = 30
)
rducks_set_execution_plan(con, plan)

reg_ipc_plus_one <- rducks_register(
  con,
  name = "r_ipc_plus_one",
  fun = function(x) x + 1L,
  args = INTEGER,
  returns = INTEGER,
  mode = "vectorized",
  side_effects = TRUE
)

dbGetQuery(con, "SELECT sum(r_ipc_plus_one(i::INTEGER)) AS x FROM range(10000) AS t(i)")
rducks_explain_udf(con, "r_ipc_plus_one")[, c(
  "name", "mode", "plan_id", "evaluator",
  "arrow_ipc_chunks", "ripc_collect_batches", "ripc_collect_requests",
  "ripc_collect_max_batch", "ripc_submit_wave_max",
  "ripc_collect_ready_max"
)]
```

For `arrow_ipc + multiprocess_parallel`, the native extension submits Arrow IPC
chunk work through the RIPC provider path and imports the returned Arrow IPC
result into DuckDB. When a RIPC callback runs on the recorded main R thread, it
can submit its own chunk and queued worker chunks before grouped collection. The
`ripc_collect_max_batch`, `ripc_submit_wave_max`, and
`ripc_collect_ready_max` counters report native queue batch/wave sizes.

### Persistent mirai workers

The experimental `ipc_mirai_pool` engine keeps worker processes alive and
preloads UDF state once per worker.

```{r mirai-ipc-plan}
mirai_plan <- rducks_execution_plan(
  "arrow_ipc", "multiprocess_parallel",
  ipc_provider = "mirai",
  ipc_workers = 1,
  ipc_max_pending = 64,
  future_timeout = 30
)
rducks_set_execution_plan(con, mirai_plan)

reg_mirai_plus_one <- rducks_register(
  con,
  name = "r_mirai_plus_one",
  fun = function(x) x + 1L,
  args = INTEGER,
  returns = INTEGER,
  mode = "vectorized",
  side_effects = TRUE
)

dbGetQuery(con, "SELECT sum(r_mirai_plus_one(i::INTEGER)) AS x FROM range(10000) AS t(i)")
rducks_explain_udf(con, "r_mirai_plus_one")[, c("engine_id", "evaluator", "arrow_ipc_chunks")]
```

A local provider-level run (`tools/benchmark_ipc_providers.R`) with one worker,
50 tasks, three iterations, and 64 rows per Arrow IPC task measured:

| provider path | median seconds | relative to `future::multisession` |
| --- | ---: | ---: |
| `future::multisession` through `ipc_future_pool` | 0.217 | 1.00x |
| `future.mirai::mirai_multisession` through `ipc_future_pool` | 0.219 | 0.99x |
| `ipc_mirai_pool`, sequential submit/collect | 0.193 | 1.12x |
| `ipc_mirai_pool`, batched submit/collect | 0.158 | 1.37x |

Those numbers are diagnostics for this machine, not a portable promise.
`future.mirai` is a useful drop-in Future backend for `ipc_future_pool`. The
separate `ipc_mirai_pool` engine exists to use Rducks' provider envelope directly:
preload evaluator/schema state once, submit only task metadata plus Arrow IPC
bytes, return structured task results, and support collect-many/collect-any style
scheduling without routing every chunk through the generic Future API. It remains
experimental while backpressure and cancellation semantics are hardened.

### Performance diagnostics

Long-running benchmarks and provider-level diagnostics live under `tools/` rather
than in the README. For cheap UDFs, `multiprocess_parallel` is often slower than
serial execution because Arrow IPC and scheduler overhead dominate the R work.
Use it when the R function is expensive enough, or when process isolation matters.

Switch back to the reference serial plan for the remaining examples.

```{r reset-after-ipc}
rducks_set_execution_plan(con, rducks_execution_plan("arrow_r", "serial"))
future::plan(future::sequential)
```

## Current scope

Rducks currently builds `rducks.duckdb_extension` at install time, loads it into
DuckDB with `rducks_enable()`, and registers scalar or vectorized R UDFs with
`rducks_register()`.

The input/output type set is `BOOLEAN`, `TINYINT`, `UTINYINT`, `SMALLINT`,
`USMALLINT`, `INTEGER`, `UINTEGER`, `BIGINT`, `UBIGINT`, `FLOAT`, `DOUBLE`,
`VARCHAR`, `BLOB`, `DATE`, `TIME`, `TIMESTAMP`, `HUGEINT`, `UHUGEINT`, `UUID`,
`INTERVAL`, `BIT`, `DECIMAL(width, scale)`, `ENUM(levels)`, and `UNION(...)`.
Composite inputs and outputs are accepted as constructed type objects such as
`TYPE[]`, `TYPE[N]`, `STRUCT(...)`, and `MAP(...)`, recursively over supported
child types.

Enum types are supported by all implemented execution plans. For
`arrow_ipc`, Rducks does not rely on general Arrow dictionary IPC because the
nanoarrow C IPC writer used by this path does not yet encode dictionary arrays.
Instead, enum payloads use the declared Rducks type descriptor as the dictionary
sidecar and serialize the DuckDB enum storage indices as ordinary Arrow integer
arrays. This is an explicit Rducks wire convention for declared `ENUM(...)`
types, not general support for arbitrary Arrow dictionary arrays. A future
same-host payload mode such as `mori` may optimize this path further.

Rducks also provides explicit R value classes for exact or DuckDB-specific
values: `rducks_bigint()`, `rducks_ubigint()`, `rducks_uuid()`,
`rducks_interval()`, `rducks_decimal()`, `rducks_hugeint()`,
`rducks_uhugeint()`, `rducks_bits()`, `rducks_enum()`, and `rducks_union()`.
Constructed DuckDB type objects are formal S7-backed descriptors with structural
validation via `rducks_is_type()`.

## Execution mode semantics

The table below is produced by `rducks_mode_semantics()`.

```{r mode-semantics, echo=FALSE, results="asis"}
knitr::kable(rducks_mode_semantics(), format = "markdown")
```

## Type descriptors

```{r type-descriptor-checks}
nested_type <- STRUCT(
  payload = UNION(code = INTEGER, label = ENUM(c("red", "blue"))),
  amount = DECIMAL(10, 2),
  tags = LIST(ENUM(c("red", "blue")))
)

rducks_is_type(nested_type)
rducks_type_kind(nested_type)
rducks_type_sql(nested_type)
rducks_type_child_names(nested_type)
rducks_check_return(UNION(code = INTEGER, label = VARCHAR), rducks_union("label", "ok"))
rducks_check_return(ENUM(c("red", "blue")), rducks_enum("red", c("red", "blue")))
```

### Argument values passed to R functions

<details>
<summary>Expand for argument values passed to R functions</summary>

The table is produced by the exported `rducks_argument_type_mapping()` helper
and reflects the currently implemented nanoarrow scalar marshalling path. With
`null_handling = "default"`, any top-level SQL `NULL` input makes DuckDB return
SQL `NULL` without calling the R function. The `SQL NULL in function` column
below applies when `null_handling = "special"`. It is type-specific: ordinary
R scalar types receive typed `NA` values, while exact/exotic value classes,
binary values, and top-level composite values receive R `NULL`. Within
homogeneous scalar lists/arrays, SQL `NULL` elements are represented as typed
`NA` values where the child type has an R `NA` representation; nested composite
`NULL` values are represented as R `NULL`.

```{r argument-type-mapping, echo=FALSE, results="asis"}
knitr::kable(rducks_argument_type_mapping(list(
  INTEGER,
  DOUBLE,
  INTEGER[],
  BIGINT[3],
  STRUCT(a = UUID, b = DECIMAL(10, 2)),
  MAP(VARCHAR, INTEGER),
  ENUM(c("red", "blue")),
  UNION(code = INTEGER, label = VARCHAR)
)), format = "markdown")
```


</details>

### NULL, NA, NaN, Inf, and value-class semantics

<details>
<summary>Expand for NULL, NA, NaN, Inf, and value-class semantics</summary>

The table below is produced by `rducks_value_semantics()`, the exported schema
that Rducks uses to document scalar-mode missing and non-finite value behavior.
Top-level R `NULL` returns map to SQL `NULL`. R `NA` values map to SQL `NULL`
when represented by the declared R type. `NaN` and `Inf` are values only for
`FLOAT` and `DOUBLE`; integer, date, time, timestamp, exact, and exotic return
paths reject them.

```{r value-semantics, echo=FALSE, results="asis"}
knitr::kable(rducks_value_semantics(list(
  INTEGER,
  DOUBLE,
  BIGINT,
  UUID,
  DECIMAL(10, 2),
  ENUM(c("red", "blue")),
  UNION(code = INTEGER, label = VARCHAR),
  MAP(VARCHAR, INTEGER)
)), format = "markdown")
```

</details>

## Composite input examples

Homogeneous scalar lists and arrays are passed as atomic R vectors with SQL
`NULL` elements represented as typed `NA` values. Structs are passed as named
lists, and maps are passed as `list(keys = ..., values = ...)`.

```{r composite-inputs}
reg_list_len <- rducks_register(
  con,
  name = "r_list_len",
  fun = function(x) length(x),
  args = INTEGER[],
  returns = INTEGER
)

reg_array_sum <- rducks_register(
  con,
  name = "r_array_sum",
  fun = function(x) sum(x),
  args = INTEGER[3],
  returns = INTEGER
)

reg_struct_sum <- rducks_register(
  con,
  name = "r_struct_sum",
  fun = function(x) x$a + x$b,
  args = STRUCT(a = INTEGER, b = INTEGER),
  returns = INTEGER
)

reg_map_sum <- rducks_register(
  con,
  name = "r_map_sum",
  fun = function(x) sum(x$values),
  args = MAP(VARCHAR, INTEGER),
  returns = INTEGER
)

reg_make_struct <- rducks_register(
  con,
  name = "r_make_struct",
  fun = function(x) list(a = x, b = x + 1L),
  args = INTEGER,
  returns = STRUCT(a = INTEGER, b = INTEGER)
)

reg_make_array <- rducks_register(
  con,
  name = "r_make_array",
  fun = function(x) c(x, x + 1L),
  args = INTEGER,
  returns = INTEGER[]
)

dbGetQuery(con, "
  SELECT
    r_list_len([1,2,3]::INTEGER[]) AS list_len,
    r_array_sum([1,2,3]::INTEGER[3]) AS array_sum,
    r_struct_sum({'a': 20, 'b': 22}::STRUCT(a INTEGER, b INTEGER)) AS struct_sum,
    r_map_sum(map(['a','b'], [20,22])) AS map_sum,
    (r_make_struct(41::INTEGER)).b AS struct_return,
    list_sum(r_make_array(20::INTEGER)) AS array_return
")
```

## NULL handling

By default, Rducks uses NULL-in/NULL-out handling: if any input is SQL `NULL`,
the R function is not called and the SQL result is `NULL`.

Use `null_handling = "special"` to pass the type-specific missing value shown
in `rducks_argument_type_mapping()` to the R function for SQL `NULL` inputs.
For ordinary scalar types this is usually a typed `NA`; for exact/exotic,
binary, and composite inputs it is R `NULL`.

```{r null-handling}
reg_null_special <- rducks_register(
  con,
  name = "r_null_special",
  fun = function(x) if (is.na(x)) 5L else x,
  args = INTEGER,
  returns = INTEGER,
  null_handling = "special"
)

dbGetQuery(con, "SELECT r_null_special(NULL::INTEGER) AS x")
```

## Exceptions and side effects

Set `exception_handling = "return_null"` to turn errors thrown by the user R
function into SQL `NULL`. Return type-checking and marshalling errors still
abort the query so type bugs are not hidden as NULLs.

```{r exception-handling}
reg_error_null <- rducks_register(
  con,
  name = "r_error_null",
  fun = function(x) stop("boom"),
  args = INTEGER,
  returns = INTEGER,
  exception_handling = "return_null"
)

dbGetQuery(con, "SELECT r_error_null(1::INTEGER) AS x")
```

Set `side_effects = TRUE` for functions with counters, randomness, I/O, or
mutation so DuckDB reruns the function for each row.

```{r side-effects}
counter <- local({
  i <- 0L
  function() {
    i <<- i + 1L
    i
  }
})

reg_counter <- rducks_register(
  con,
  name = "r_counter",
  fun = counter,
  args = character(),
  returns = INTEGER,
  side_effects = TRUE
)

dbGetQuery(con, "SELECT r_counter() AS x FROM range(5)")

set.seed(1)
reg_rng <- rducks_register(
  con,
  name = "r_rng",
  fun = function() runif(1),
  args = character(),
  returns = DOUBLE,
  side_effects = TRUE
)

dbGetQuery(con, "SELECT r_rng() AS x FROM range(3)")
```

## Build notes

The package builds its DuckDB extension during installation using `configure` or
`configure.win`. The extension metadata footer is appended by
`tools/append_extension_metadata.R`.

DuckDB C API headers are refreshed explicitly with:

```sh
Rscript tools/fetch_duckdb_headers.R --ref v1.5.2
```

The DuckDB Arrow C Data path requires DuckDB's unstable C extension API, so the
extension metadata uses `C_STRUCT_UNSTABLE` and must match the bundled DuckDB
header/runtime version. See `docs/BUILD.md` for the extension build and metadata
details.

## DuckDB C extension ABI surface

The extension entrypoint receives the DuckDB database handle from DuckDB's
extension access struct, stores it in a per-database runtime record, and opens an
extension-owned DuckDB connection for native callback support. On extension
reload, Rducks verifies that the active connection has the SQL surface registered
and re-registers it when needed.

The unstable DuckDB C extension ABI entries are tracked by
`tools/used_duckdb_unstable_api.R` and documented in `docs/BUILD.md`. The README
intentionally does not source repository helper scripts while rendering. The
current output of
`source("tools/used_duckdb_unstable_api.R"); cat(rducks_used_duckdb_unstable_api_markdown("."))`
is:

| ABI group | Functions used | Count |
| --- | --- | ---: |
| `unstable_new_arrow_functions` | `duckdb_data_chunk_from_arrow`, `duckdb_data_chunk_to_arrow`, `duckdb_destroy_arrow_converted_schema`, `duckdb_schema_from_arrow`, `duckdb_to_arrow_schema` | 5 |
| `unstable_new_error_data_functions` | `duckdb_destroy_error_data`, `duckdb_error_data_has_error`, `duckdb_error_data_message` | 3 |
| `unstable_new_open_connect_functions` | `duckdb_client_context_get_connection_id`, `duckdb_connection_get_arrow_options`, `duckdb_destroy_arrow_options`, `duckdb_destroy_client_context` | 4 |
| `unstable_new_scalar_function_functions` | `duckdb_scalar_function_bind_get_extra_info`, `duckdb_scalar_function_bind_set_error`, `duckdb_scalar_function_get_client_context`, `duckdb_scalar_function_set_bind`, `duckdb_scalar_function_set_bind_data`, `duckdb_scalar_function_set_bind_data_copy` | 6 |
| `unstable_new_scalar_function_state_functions` | `duckdb_scalar_function_get_state`, `duckdb_scalar_function_init_get_bind_data`, `duckdb_scalar_function_init_get_client_context`, `duckdb_scalar_function_init_get_extra_info`, `duckdb_scalar_function_init_set_error`, `duckdb_scalar_function_init_set_state`, `duckdb_scalar_function_set_init` | 7 |
| `unstable_new_vector_functions` | `duckdb_create_selection_vector`, `duckdb_destroy_selection_vector`, `duckdb_selection_vector_get_data_ptr`, `duckdb_vector_copy_sel` | 4 |
