---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
library(Rducks)
library(future)
library(future.mirai)
library(nanoarrow)
```

# Rducks

[![R-CMD-check](https://github.com/sounkou-bioinfo/Rducks/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rducks/actions/workflows/R-CMD-check.yaml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Rducks)](https://sounkou-bioinfo.r-universe.dev/Rducks)

Rducks registers R functions as DuckDB SQL functions. It ships as an R package
plus a DuckDB extension. UDF inputs and outputs move through explicit execution
plans:

- `arrow_r`: reference path using DuckDB Arrow C Data plus nanoarrow/R.
- `arrow_c`: native extension path for supported scalar and vectorized calls.
- `arrow_ipc`: process-isolated path using Arrow IPC request/result payloads
  and a generic Future backend.

The user-facing UDF semantics are separate from the execution plan:
`mode = "scalar"` means one R call per logical row; `mode = "vectorized"` means
one R call per DuckDB chunk.

## Quick start

```{r getting-started}
library(DBI)
library(duckdb)
library(Rducks)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(con, threads = "single")

reg_plus_one <- rducks_register(
  con,
  name = "r_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "scalar"
)

dbGetQuery(con, "SELECT r_plus_one(41.0) AS x")
```

`rducks_register()` returns an `rducks_registration` object that records the
connection, normalized signature, and registration options. You do not need to
keep this object for the UDF to keep working in DuckDB.

## Scalar and vectorized modes

Scalar mode calls the R function once per DuckDB row. Vectorized mode calls the
R function once per DuckDB chunk with one R vector/list-column per declared
argument.

```{r vectorized-mode}
reg_vec_plus_one <- rducks_register(
  con,
  name = "r_vec_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "vectorized",
  side_effects = TRUE
)

dbGetQuery(con, "SELECT sum(r_vec_plus_one(i::DOUBLE)) AS x FROM range(5) AS t(i)")
```

In vectorized mode, `null_handling = "default"` evaluates only rows with no
top-level SQL NULL inputs and scatters SQL NULLs back into the result.
`null_handling = "special"` passes all rows using the same NA/NULL shapes as
scalar mode. The return length must match the number of evaluated rows.
Vectorized mode currently requires at least one declared argument.

A tiny benchmark with `bench` can show the call-shape difference for simple R
work. The result is illustrative rather than a performance guarantee.

```{r vectorized-benchmark}
bench_n <- 10000L
bench_result <- bench::mark(
  scalar = DBI::dbGetQuery(
    con,
    sprintf("SELECT sum(r_plus_one(i::DOUBLE)) AS x FROM range(%d) AS t(i)", bench_n)
  ),
  vectorized = DBI::dbGetQuery(
    con,
    sprintf("SELECT sum(r_vec_plus_one(i::DOUBLE)) AS x FROM range(%d) AS t(i)", bench_n)
  ),
  iterations = 3,
  check = FALSE
)
bench_result[, c("expression", "median", "itr/sec", "mem_alloc")]
```

## Execution plans

Execution plans choose the marshalling implementation and concurrency contract
for a connection. Registration remains semantic: name, function, mode, types,
NULL handling, exception handling, and side-effect flag.

| Plan | Scalar | Vectorized | Notes |
| --- | --- | --- | --- |
| `arrow_r + serial` | implemented | implemented | reference implementation |
| `arrow_r + inproc_concurrent` | implemented | implemented | queued same-process callbacks; R API work stays on the recorded R lane |
| `arrow_c + serial` | implemented | implemented | native evaluator token `RC` for scalar, `RCV` for vectorized |
| `arrow_c + inproc_concurrent` | implemented | implemented | queued same-process callbacks with `arrow_c` marshalling |
| `arrow_ipc + multiprocess_parallel` | implemented | implemented | Arrow IPC request/result payloads through the active Future backend; evaluator token `RIPC` |

`arrow_r + serial` is the semantic reference. Other implemented plans are tested
against it and do not silently fall back to another marshalling path. Use
`rducks_explain_udf()` to inspect the plan and native counters for a registered
UDF.

```{r explain-udf}
rducks_explain_udf(con, "r_vec_plus_one")[, c(
  "name", "mode", "plan_id", "native_marshalling",
  "evaluator", "arrow_r_chunks", "arrow_c_chunks", "arrow_ipc_chunks"
)]
```

## In-process queued execution

Register UDFs in the registration-safe configuration. Then call
`rducks_enable_inproc()` to switch query execution to the explicit in-process
queue. This is a liveness and scheduling feature for same-process callbacks; it
still serializes R API work on the recorded R lane.

DuckDB interprets `external_threads` as threads supplied by the caller rather
than DuckDB-created worker threads. For stress tests, prefer settings such as
`threads = 4, external_threads = 1`; do not set `external_threads = threads`
unless you intentionally want no DuckDB-created worker pool.

```{r inproc-queue}
reg_sleepy <- rducks_register(
  con,
  name = "r_sleepy_time",
  fun = function(x) {
    Sys.sleep(0.001)
    Sys.time()
  },
  args = DOUBLE,
  returns = TIMESTAMP,
  side_effects = TRUE
)

rducks_enable_inproc(con)

rducks_inproc_self_test(con, 3)
rducks_inproc_stats(con)

dbGetQuery(con, "SELECT r_sleepy_time(1.0) AS x")
rducks_inproc_stats(con)

rducks_disable_inproc(con, threads = 1)
```

## Multiprocess Arrow IPC execution

`arrow_ipc + multiprocess_parallel` uses Arrow IPC bytes for chunk tasks and
results. Scalar mode loops over rows inside the worker process; vectorized mode
calls the R function once per chunk inside the worker process. Configure the
Future backend before executing queries.

```{r arrow-ipc-plan}
future::plan(future.mirai::mirai_multisession, workers = 2)

plan <- rducks_execution_plan(
  "arrow_ipc", "multiprocess_parallel",
  future_packages = "stats",
  future_timeout = 30
)
rducks_set_execution_plan(con, plan)

reg_ipc_plus_one <- rducks_register(
  con,
  name = "r_ipc_plus_one",
  fun = function(x) x + 1L,
  args = INTEGER,
  returns = INTEGER,
  mode = "vectorized",
  side_effects = TRUE
)

dbGetQuery(con, "SELECT sum(r_ipc_plus_one(i::INTEGER)) AS x FROM range(10000) AS t(i)")
rducks_explain_udf(con, "r_ipc_plus_one")[, c(
  "name", "mode", "plan_id", "evaluator",
  "arrow_ipc_chunks", "ripc_collect_batches", "ripc_collect_requests"
)]
```

The owned-task shape is what allows real worker parallelism: encode every input
chunk to an owned Arrow IPC payload, submit all tasks, then collect owned Arrow
IPC result payloads.

```{r owned-ipc-pipeline-benchmark}
workers <- 4L
n <- 8192L
chunk_size <- 1024L
sleep_s <- 0.1

# Benchmark-only Future polling tweak: small worker tasks can finish before
# the default polling interval notices them. This reduces wait latency; it is
# not required for Rducks correctness or for real parallelism.
options(future.wait.interval = 0.001, future.wait.alpha = 1.01)
future::plan(future.mirai::mirai_multisession, workers = workers)
# Warm workers so the measurement reflects chunk scheduling, not startup.
invisible(future::value(future::future(NULL)))

rows <- 0:(n - 1L)
chunks <- split(rows, ceiling(seq_along(rows) / chunk_size))
ns <- lengths(chunks)

encode_input <- function(x) {
  Rducks:::rducks_arrow_ipc_encode(
    nanoarrow::as_nanoarrow_array(data.frame(arg1 = as.integer(x)))
  )
}

output_schema_spec <- Rducks:::rducks_arrow_schema_to_spec(
  nanoarrow::infer_nanoarrow_schema(data.frame(result = integer()))
)

worker_eval <- function(payload, n_rows) {
  Rducks:::rducks_future_worker_eval_arrow_ipc_chunk(
    input_payload = payload,
    output_schema_spec = output_schema_spec,
    n = n_rows,
    fun = function(x) {
      Sys.sleep(sleep_s)
      x + 1L
    },
    arg_types = list(INTEGER),
    return_type = INTEGER,
    null_handling = "default",
    exception_handling = "rethrow",
    mode = "vectorized"
  )
}

payloads <- lapply(chunks, encode_input)

seq_time <- system.time({
  seq_payloads <- Map(worker_eval, payloads, ns)
})[["elapsed"]]

par_time <- system.time({
  futures <- Map(function(payload, n_rows) {
    future::future(
      worker_eval(payload, n_rows),
      globals = list(worker_eval = worker_eval, payload = payload, n_rows = n_rows),
      packages = "Rducks",
      stdout = FALSE
    )
  }, payloads, ns)
  par_payloads <- future::value(futures, stdout = FALSE, signal = FALSE)
})[["elapsed"]]

data.frame(
  rows = n,
  chunks = length(chunks),
  workers = workers,
  sequential_s = seq_time,
  parallel_s = par_time,
  speedup = seq_time / par_time
)
```

The same benchmark is available as `tools/benchmark_owned_ipc_pipeline.R` for
repeatable source-checkout runs. Results depend on provider, workload, chunk
size, and worker startup state.

## Current scope

Rducks currently builds `rducks.duckdb_extension` at install time, loads it into
DuckDB with `rducks_enable()`, and registers scalar or vectorized R UDFs with
`rducks_register()`.

The input/output type set is `BOOLEAN`, `TINYINT`, `UTINYINT`, `SMALLINT`,
`USMALLINT`, `INTEGER`, `UINTEGER`, `BIGINT`, `UBIGINT`, `FLOAT`, `DOUBLE`,
`VARCHAR`, `BLOB`, `DATE`, `TIME`, `TIMESTAMP`, `HUGEINT`, `UHUGEINT`, `UUID`,
`INTERVAL`, `BIT`, `DECIMAL(width, scale)`, `ENUM(levels)`, and `UNION(...)`.
Composite inputs and outputs are accepted as constructed type objects such as
`TYPE[]`, `TYPE[N]`, `STRUCT(...)`, and `MAP(...)`, recursively over supported
child types.

Enum types are supported by all implemented execution plans. For
`arrow_ipc`, Rducks does not rely on general Arrow dictionary IPC because the
nanoarrow C IPC writer used by this path does not yet encode dictionary arrays.
Instead, enum payloads use the declared Rducks type descriptor as the dictionary
sidecar and serialize the DuckDB enum storage indices as ordinary Arrow integer
arrays. This is an explicit Rducks wire convention for declared `ENUM(...)`
types, not general support for arbitrary Arrow dictionary arrays. A future
same-host payload mode such as `mori` may optimize this path further.

Rducks also provides explicit R value classes for exact or DuckDB-specific
values: `rducks_bigint()`, `rducks_ubigint()`, `rducks_uuid()`,
`rducks_interval()`, `rducks_decimal()`, `rducks_hugeint()`,
`rducks_uhugeint()`, `rducks_bits()`, `rducks_enum()`, and `rducks_union()`.
Constructed DuckDB type objects are formal S7-backed descriptors with structural
validation via `rducks_is_type()`.

## Execution mode semantics

The table below is produced by `rducks_mode_semantics()`.

```{r mode-semantics, echo=FALSE, results="asis"}
mode_semantics <- Rducks::rducks_mode_semantics()
knitr::kable(
  data.frame(
    mode = paste0("`", mode_semantics$mode, "`"),
    status = mode_semantics$status,
    call_granularity = mode_semantics$call_granularity,
    input_shape = mode_semantics$input_shape,
    return_shape = mode_semantics$return_shape,
    length_semantics = mode_semantics$length_semantics,
    threading = mode_semantics$threading,
    copy_semantics = mode_semantics$copy_semantics,
    check.names = FALSE
  ),
  format = "markdown"
)
```

## Type descriptors

```{r type-descriptor-checks}
nested_type <- STRUCT(
  payload = UNION(code = INTEGER, label = ENUM(c("red", "blue"))),
  amount = DECIMAL(10, 2),
  tags = LIST(ENUM(c("red", "blue")))
)

rducks_is_type(nested_type)
rducks_type_kind(nested_type)
rducks_type_sql(nested_type)
rducks_type_child_names(nested_type)
rducks_check_return(UNION(code = INTEGER, label = VARCHAR), rducks_union("label", "ok"))
rducks_check_return(ENUM(c("red", "blue")), rducks_enum("red", c("red", "blue")))
```

### Argument values passed to R functions

<details>
<summary>Expand for argument values passed to R functions</summary>

The table is produced by the exported `rducks_argument_type_mapping()` helper
and reflects the currently implemented nanoarrow scalar marshalling path. With
`null_handling = "default"`, any top-level SQL `NULL` input makes DuckDB return
SQL `NULL` without calling the R function. The `SQL NULL in function` column
below applies when `null_handling = "special"`. It is type-specific: ordinary
R scalar types receive typed `NA` values, while exact/exotic value classes,
binary values, and top-level composite values receive R `NULL`. Within
homogeneous scalar lists/arrays, SQL `NULL` elements are represented as typed
`NA` values where the child type has an R `NA` representation; nested composite
`NULL` values are represented as R `NULL`.

```{r argument-type-mapping, echo=FALSE, results="asis"}
argument_type_mapping <- rbind(
  Rducks::rducks_argument_type_mapping(),
  Rducks::rducks_argument_type_mapping(list(
    INTEGER[],
    BIGINT[3],
    STRUCT(a = UUID, b = DECIMAL(10, 2)),
    MAP(VARCHAR, INTEGER),
    DECIMAL(10, 2),
    ENUM(c("red", "blue")),
    UNION(code = INTEGER, label = VARCHAR)
  ))
)
knitr::kable(
  data.frame(
    argument_type = paste0("`", argument_type_mapping$duckdb_sql, "`"),
    r_type = argument_type_mapping$r_type,
    r_value_passed_to_fun = argument_type_mapping$r_value_passed_to_fun,
    sql_null_in_function = argument_type_mapping$sql_null_in_function,
    copy_semantics = argument_type_mapping$copy_semantics,
    notes = argument_type_mapping$notes,
    check.names = FALSE
  ),
  format = "markdown"
)
```


</details>

### NULL, NA, NaN, Inf, and value-class semantics

<details>
<summary>Expand for NULL, NA, NaN, Inf, and value-class semantics</summary>

The table below is produced by `rducks_value_semantics()`, the exported schema
that Rducks uses to document scalar-mode missing and non-finite value behavior.
Top-level R `NULL` returns map to SQL `NULL`. R `NA` values map to SQL `NULL`
when represented by the declared R type. `NaN` and `Inf` are values only for
`FLOAT` and `DOUBLE`; integer, date, time, timestamp, exact, and exotic return
paths reject them.

```{r value-semantics, echo=FALSE, results="asis"}
value_semantics <- Rducks::rducks_value_semantics(list(
  INTEGER,
  DOUBLE,
  BIGINT,
  UBIGINT,
  HUGEINT,
  UHUGEINT,
  UUID,
  INTERVAL,
  BIT,
  DECIMAL(10, 2),
  ENUM(c("red", "blue")),
  UNION(code = INTEGER, label = VARCHAR),
  STRUCT(amount = DECIMAL(10, 2), id = BIGINT),
  MAP(VARCHAR, INTEGER)
))
knitr::kable(
  data.frame(
    duckdb_type = paste0("`", value_semantics$duckdb_sql, "`"),
    sql_null_special = value_semantics$sql_null_input_special,
    r_na_return = value_semantics$r_na_return,
    r_nan_return = value_semantics$r_nan_return,
    r_inf_return = value_semantics$r_inf_return,
    binary_ops = value_semantics$binary_ops,
    error_semantics = value_semantics$error_semantics,
    check.names = FALSE
  ),
  format = "markdown"
)
```

</details>

## Composite input examples

Homogeneous scalar lists and arrays are passed as atomic R vectors with SQL
`NULL` elements represented as typed `NA` values. Structs are passed as named
lists, and maps are passed as `list(keys = ..., values = ...)`.

```{r composite-inputs}
reg_list_len <- rducks_register(
  con,
  name = "r_list_len",
  fun = function(x) length(x),
  args = INTEGER[],
  returns = INTEGER
)

reg_array_sum <- rducks_register(
  con,
  name = "r_array_sum",
  fun = function(x) sum(x),
  args = INTEGER[3],
  returns = INTEGER
)

reg_struct_sum <- rducks_register(
  con,
  name = "r_struct_sum",
  fun = function(x) x$a + x$b,
  args = STRUCT(a = INTEGER, b = INTEGER),
  returns = INTEGER
)

reg_map_sum <- rducks_register(
  con,
  name = "r_map_sum",
  fun = function(x) sum(x$values),
  args = MAP(VARCHAR, INTEGER),
  returns = INTEGER
)

reg_make_struct <- rducks_register(
  con,
  name = "r_make_struct",
  fun = function(x) list(a = x, b = x + 1L),
  args = INTEGER,
  returns = STRUCT(a = INTEGER, b = INTEGER)
)

reg_make_array <- rducks_register(
  con,
  name = "r_make_array",
  fun = function(x) c(x, x + 1L),
  args = INTEGER,
  returns = INTEGER[]
)

dbGetQuery(con, paste(
  "SELECT",
  "r_list_len([1,2,3]::INTEGER[]) AS list_len,",
  "r_array_sum([1,2,3]::INTEGER[3]) AS array_sum,",
  "r_struct_sum({'a': 20, 'b': 22}::STRUCT(a INTEGER, b INTEGER)) AS struct_sum,",
  "r_map_sum(map(['a','b'], [20,22])) AS map_sum,",
  "(r_make_struct(41::INTEGER)).b AS struct_return,",
  "list_sum(r_make_array(20::INTEGER)) AS array_return"
))
```

## NULL handling

By default, Rducks uses NULL-in/NULL-out handling: if any input is SQL `NULL`,
the R function is not called and the SQL result is `NULL`.

Use `null_handling = "special"` to pass the type-specific missing value shown
in `rducks_argument_type_mapping()` to the R function for SQL `NULL` inputs.
For ordinary scalar types this is usually a typed `NA`; for exact/exotic,
binary, and composite inputs it is R `NULL`.

```{r null-handling}
reg_null_special <- rducks_register(
  con,
  name = "r_null_special",
  fun = function(x) if (is.na(x)) 5L else x,
  args = INTEGER,
  returns = INTEGER,
  null_handling = "special"
)

dbGetQuery(con, "SELECT r_null_special(NULL::INTEGER) AS x")
```

## Exceptions and side effects

Set `exception_handling = "return_null"` to turn R errors into SQL
`NULL`.

```{r exception-handling}
reg_error_null <- rducks_register(
  con,
  name = "r_error_null",
  fun = function(x) stop("boom"),
  args = INTEGER,
  returns = INTEGER,
  exception_handling = "return_null"
)

dbGetQuery(con, "SELECT r_error_null(1::INTEGER) AS x")
```

Set `side_effects = TRUE` for functions with counters, randomness, I/O, or
mutation so DuckDB reruns the function for each row.

```{r side-effects}
counter <- local({
  i <- 0L
  function() {
    i <<- i + 1L
    i
  }
})

reg_counter <- rducks_register(
  con,
  name = "r_counter",
  fun = counter,
  args = character(),
  returns = INTEGER,
  side_effects = TRUE
)

dbGetQuery(con, "SELECT r_counter() AS x FROM range(5)")

set.seed(1)
reg_rng <- rducks_register(
  con,
  name = "r_rng",
  fun = function() runif(1),
  args = character(),
  returns = DOUBLE,
  side_effects = TRUE
)

dbGetQuery(con, "SELECT r_rng() AS x FROM range(3)")
```

## Build notes

The package builds its DuckDB extension during installation using `configure` or
`configure.win`. The extension metadata footer is appended by
`tools/append_extension_metadata.R`.

DuckDB C API headers are refreshed explicitly with:

```sh
Rscript tools/fetch_duckdb_headers.R --ref v1.5.2
```

The DuckDB Arrow C Data path requires DuckDB's unstable C extension API, so the
extension metadata uses `C_STRUCT_UNSTABLE` and must match the bundled DuckDB header/runtime
version. See `docs/BUILD.md` for the extension build and metadata details.

```{r cleanup, include=FALSE}
dbDisconnect(con, shutdown = TRUE)
```
