---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Rducks

Rducks registers R scalar functions as DuckDB SQL functions. It ships as an R
package plus a DuckDB extension. Rtinycc compiles a small C wrapper for each R
function signature, and the loaded DuckDB extension registers that wrapper on a
DuckDB connection.

## Current scope

Rducks currently builds `rducks.duckdb_extension` at install time, loads it into
DuckDB with `rducks_enable()`, and registers row-oriented scalar R UDFs with
`rducks_register()`. The implemented scalar input/output type set is `bool`,
`i8`, `u8`, `i16`, `u16`, `i32`, `u32`, `i64`, `u64`, `f32`, `f64`,
`varchar`, `blob`, `date`, `time`, and `timestamp`. Composite inputs are
accepted as `list<type>`, `type[]`, `type[N]`, `struct<name:type;...>`, and
`map<key;value>`. Registration also supports `null_handling`,
`exception_handling`, and `side_effects` controls.

Composite return types are not implemented yet. Direct R callbacks require
single-thread DuckDB execution, so call `rducks_enable(con, threads = "single")`
or set `PRAGMA threads=1` before registering R UDFs.

## Example

```{r example}
library(DBI)
library(duckdb)
library(Rducks)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(con, threads = "single")

reg_plus_one <- rducks_register(
  con,
  name = "r_plus_one",
  fun = function(x) x + 1,
  args = "f64",
  returns = "f64"
)

dbGetQuery(con, "SELECT r_plus_one(41.0) AS x")
```

`u32`, `i64`, and `u64` are passed through R numeric (`double`), matching the
DuckDB R package's default type mapping. Values beyond `2^53` cannot be exactly
represented as R doubles.

<details>
<summary>Argument values passed to R callbacks</summary>

The table is produced by the exported `rducks_argument_type_mapping()` helper.
With `null_handling = "default"`, any top-level SQL `NULL` input makes DuckDB
return SQL `NULL` without calling the R callback. The `SQL NULL in callback`
column below applies when `null_handling = "special"`. Nested SQL `NULL` values
inside composite inputs are represented as R `NULL`.

```{r argument-type-mapping, echo=FALSE, comment="", R.options=list(width=220)}
type_mapping_tokens <- c(
  "bool", "i8", "u8", "i16", "u16", "i32", "u32", "i64", "u64",
  "f32", "f64", "varchar", "blob", "date", "time", "timestamp",
  "list<i32>", "i32[]", "i32[3]", "struct<a:i32;b:varchar>",
  "map<varchar;i32>"
)
argument_type_mapping <- rducks_argument_type_mapping(type_mapping_tokens)
print(
  argument_type_mapping[, c(
    "rducks_type", "duckdb_sql", "r_value_passed_to_fun",
    "sql_null_in_callback", "notes"
  )],
  row.names = FALSE,
  right = FALSE
)
```

</details>

## NULL handling

By default, Rducks uses NULL-in/NULL-out handling: if any input is SQL `NULL`,
the R callback is not called and the SQL result is `NULL`.

Use `null_handling = "special"` to pass an R NA-like value to the R function
for SQL `NULL` inputs.

```{r null-handling}
reg_null_special <- rducks_register(
  con,
  name = "r_null_special",
  fun = function(x) if (is.na(x)) 5L else x,
  args = "i32",
  returns = "i32",
  null_handling = "special"
)

dbGetQuery(con, "SELECT r_null_special(NULL::INTEGER) AS x")
```

## Exceptions and side effects

Set `exception_handling = "return_null"` to turn callback errors into SQL
`NULL`.

```{r exception-handling}
reg_error_null <- rducks_register(
  con,
  name = "r_error_null",
  fun = function(x) stop("boom"),
  args = "i32",
  returns = "i32",
  exception_handling = "return_null"
)

dbGetQuery(con, "SELECT r_error_null(1::INTEGER) AS x")
```

Set `side_effects = TRUE` for callbacks with counters, randomness, I/O, or
mutation so DuckDB reruns the function for each row.

```{r side-effects}
counter <- local({
  i <- 0L
  function() {
    i <<- i + 1L
    i
  }
})

reg_counter <- rducks_register(
  con,
  name = "r_counter",
  fun = counter,
  args = character(),
  returns = "i32",
  side_effects = TRUE
)

dbGetQuery(con, "SELECT r_counter() AS x FROM range(5)")
```

## Build notes

The package builds its DuckDB extension during installation using `configure` or
`configure.win`. The extension metadata footer is appended by
`tools/append_extension_metadata.R`.

DuckDB C API headers are refreshed explicitly with:

```sh
Rscript tools/fetch_duckdb_headers.R --ref v1.2.0
```

See `docs/BUILD.md` for the extension build and metadata details.

```{r cleanup, include=FALSE}
dbDisconnect(con, shutdown = TRUE)
```
