---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
library(Rducks)
library(future)
library(future.mirai)
library(nanoarrow)
```

# Rducks

[![R-CMD-check](https://github.com/sounkou-bioinfo/Rducks/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rducks/actions/workflows/R-CMD-check.yaml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Rducks)](https://sounkou-bioinfo.r-universe.dev/Rducks)

Rducks registers R functions as DuckDB SQL functions using a DuckDB C
extension, including a small set of unstable DuckDB C extension API functions.
The extension records the DuckDB database instance handle at initialization and
keeps an extension-owned connection associated with that runtime. UDF inputs and
outputs move through explicit execution plans:

- `arrow_r`: reference path using DuckDB Arrow C Data plus nanoarrow/R.
- `arrow_c`: native extension path for supported scalar calls with direct
  DuckDB-vector materialization. Vectorized `arrow_c` is intentionally not
  exposed until it has a direct implementation rather than the old Arrow/R
  helper bridge.
- `arrow_ipc`: process-isolated R execution using Arrow IPC request/result
  payloads and a generic Future backend.

The user-facing UDF semantics are separate from the execution plan:
`mode = "scalar"` means one R call per logical row; `mode = "vectorized"` means
one R call per DuckDB chunk.

## Quick start

```{r getting-started}
library(DBI)
library(duckdb)
library(Rducks)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(con, threads = "single")

reg_hello <- rducks_register(
  con,
  name = "r_hello",
  fun = function() "hello from R",
  args = NULL,
  returns = VARCHAR,
  mode = "scalar"
)

dbGetQuery(con, "SELECT r_hello() AS message")

reg_plus_one <- rducks_register(
  con,
  name = "r_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "scalar"
)

dbGetQuery(con, "SELECT r_plus_one(41.0) AS x")
```

Use `args = NULL` for a zero-argument scalar UDF. `null_handling` only affects
SQL NULL inputs, so the default is appropriate for `r_hello()`.

`rducks_register()` returns an `rducks_registration` object that records the
connection, normalized signature, and registration options. You do not need to
keep this object for the UDF to keep working in DuckDB.

### Arrow conversion setting

`rducks_enable()` sets DuckDB's `arrow_lossless_conversion=true` setting on the
user connection, and the Rducks extension also sets it on its own internal
DuckDB connection. Rducks uses DuckDB's Arrow C Data conversion APIs for several
execution plans, and this setting tells DuckDB to preserve DuckDB-specific type
identity in Arrow metadata when possible instead of exporting only a more generic
Arrow representation. This matters for types such as `UUID`, `BIT`,
`HUGEINT`/`UHUGEINT`, `TIME WITH TIME ZONE`, and JSON aliases. It does not
change stored DuckDB values or SQL evaluation; it only affects DuckDB-to-Arrow
conversion used at the Rducks boundary.

## Scalar and vectorized modes

Scalar mode calls the R function once per DuckDB row. Vectorized mode calls the
R function once per DuckDB chunk with one R vector/list-column per declared
argument.

```{r vectorized-mode}
reg_vec_plus_one <- rducks_register(
  con,
  name = "r_vec_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "vectorized",
  side_effects = TRUE
)

dbGetQuery(con, "SELECT sum(r_vec_plus_one(i::DOUBLE)) AS x FROM range(5) AS t(i)")
```

In vectorized mode, `null_handling = "default"` evaluates only rows with no
top-level SQL NULL inputs and scatters SQL NULLs back into the result.
`null_handling = "special"` passes all rows using the same NA/NULL shapes as
scalar mode. The return length must match the number of evaluated rows.
Vectorized mode currently requires at least one declared argument.

A tiny benchmark with `bench` can show the call-shape difference for simple R
work. The result is illustrative rather than a performance guarantee.

```{r vectorized-benchmark}
bench_n <- 10000L
bench_result <- bench::mark(
  scalar = DBI::dbGetQuery(
    con,
    sprintf("SELECT sum(r_plus_one(i::DOUBLE)) AS x FROM range(%d) AS t(i)", bench_n)
  ),
  vectorized = DBI::dbGetQuery(
    con,
    sprintf("SELECT sum(r_vec_plus_one(i::DOUBLE)) AS x FROM range(%d) AS t(i)", bench_n)
  ),
  iterations = 3,
  check = FALSE
)
bench_result[, c("expression", "median", "itr/sec", "mem_alloc")]
```

## Execution plans

Execution plans choose the marshalling implementation and concurrency contract
for a connection. Registration remains semantic: name, function, mode, types,
NULL handling, exception handling, and side-effect flag.

| Plan | Scalar | Vectorized | Notes |
| --- | --- | --- | --- |
| `arrow_r + serial` | implemented | implemented | reference implementation |
| `arrow_r + inproc_concurrent` | implemented | implemented | queued same-process callbacks; R API work stays on the recorded main R thread |
| `arrow_c + serial` | implemented | not implemented | direct native evaluator token `RC` for scalar |
| `arrow_c + inproc_concurrent` | implemented | not implemented | queued same-process callbacks with scalar `arrow_c` marshalling |
| `arrow_ipc + multiprocess_parallel` | implemented | implemented | Arrow IPC request/result payloads through the active Future backend; evaluator token `RIPC` |

`arrow_r + serial` is the semantic reference. Other implemented plans are tested
against it and do not silently fall back to another marshalling path. Use
`rducks_explain_udf()` to inspect the plan and native counters for a registered
UDF.

```{r explain-udf}
rducks_explain_udf(con, "r_vec_plus_one")[, c(
  "name", "mode", "plan_id", "native_marshalling",
  "evaluator", "arrow_r_chunks", "arrow_c_chunks", "arrow_ipc_chunks"
)]
```

## In-process queued execution

Register UDFs in the registration-safe configuration. Then call
`rducks_enable_inproc()` to switch query execution to the explicit in-process
queue. This is a liveness and scheduling feature for same-process callbacks; it
still serializes R API work on the recorded main R thread.

DuckDB interprets `external_threads` as threads supplied by the caller rather
than DuckDB-created worker threads. For stress tests, prefer settings such as
`threads = 4, external_threads = 1`; do not set `external_threads = threads`
unless you intentionally want no DuckDB-created worker pool.

The example below uses a DuckDB table scan, not an Rducks table source. It is a
diagnostic, not a speed benchmark. The `rducks_thread_is_main()` probe shows
whether DuckDB evaluated a scan on the recorded main R thread or on DuckDB
worker threads. The UDF query then runs over the same table and the queue
counters show how many DuckDB chunks were routed through the in-process queue.
Do not infer an expected speed gain from this mode: R API work is still
serialized on the recorded main R thread, and the queue plus Arrow marshalling
add overhead.

```{r inproc-queue}
reg_inproc_plus_one <- rducks_register(
  con,
  name = "r_inproc_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE
)

rducks_enable_inproc(con, threads = 4, external_threads = 1)
rducks_current_execution_plan(con)$plan_id

dbGetQuery(
  con,
  "SELECT current_setting('threads') AS threads, current_setting('external_threads') AS external_threads"
)

inproc_n <- 200000L
invisible(DBI::dbExecute(
  con,
  sprintf(
    "CREATE TEMP TABLE inproc_input AS SELECT i::DOUBLE AS x, i::UBIGINT AS tid FROM range(%d) AS t(i)",
    inproc_n
  )
))

dbGetQuery(
  con,
  paste(
    "SELECT count(*)::VARCHAR AS rows,",
    "sum(CASE WHEN rducks_thread_is_main(tid) THEN 1 ELSE 0 END)::VARCHAR AS main_thread_rows,",
    "(count(*) - sum(CASE WHEN rducks_thread_is_main(tid) THEN 1 ELSE 0 END))::VARCHAR AS worker_thread_rows",
    "FROM inproc_input"
  )
)

stats0 <- rducks_inproc_stats(con)
dbGetQuery(con, "SELECT sum(r_inproc_plus_one(x)) AS total FROM inproc_input")
stats1 <- rducks_inproc_stats(con)
stats1 - stats0

rducks_disable_inproc(con, threads = 1)
```

## Multiprocess Arrow IPC execution

`arrow_ipc + multiprocess_parallel` uses Arrow IPC bytes for chunk tasks and
results. Scalar mode loops over rows inside the worker process; vectorized mode
calls the R function once per chunk inside the worker process. Configure the
Future backend before executing queries.

```{r arrow-ipc-plan}
future::plan(future.mirai::mirai_multisession, workers = 2)

plan <- rducks_execution_plan(
  "arrow_ipc", "multiprocess_parallel",
  future_packages = "stats",
  future_timeout = 30
)
rducks_set_execution_plan(con, plan)

reg_ipc_plus_one <- rducks_register(
  con,
  name = "r_ipc_plus_one",
  fun = function(x) x + 1L,
  args = INTEGER,
  returns = INTEGER,
  mode = "vectorized",
  side_effects = TRUE
)

dbGetQuery(con, "SELECT sum(r_ipc_plus_one(i::INTEGER)) AS x FROM range(10000) AS t(i)")
rducks_explain_udf(con, "r_ipc_plus_one")[, c(
  "name", "mode", "plan_id", "evaluator",
  "arrow_ipc_chunks", "ripc_collect_batches", "ripc_collect_requests",
  "ripc_collect_max_batch", "ripc_submit_wave_max",
  "ripc_collect_ready_max"
)]
```

For `arrow_ipc + multiprocess_parallel`, the native extension submits Arrow IPC
chunk work through the Future-backed RIPC path and imports the returned Arrow IPC
result into DuckDB. When a RIPC callback runs on the recorded main R thread, it can
submit its own chunk and queued worker chunks before grouped collection. The
`ripc_collect_max_batch`, `ripc_submit_wave_max`, and
`ripc_collect_ready_max` counters report native queue batch/wave sizes.

### DuckDB table-scan no-op benchmark

This benchmark uses a DuckDB table scan over a table created from `range()`, not
an Rducks table source. The `parallel_probe` query verifies that DuckDB actually
uses a non-main worker thread for this table scan before timing the no-op UDF.

```{r arrow-ipc-table-noop-benchmark}
arrow_ipc_noop_benchmark <- local({
  future::plan(future.mirai::mirai_multisession, workers = 4)

  bench_con <- DBI::dbConnect(
    duckdb::duckdb(config = list(allow_unsigned_extensions = "true")),
    dbdir = ":memory:"
  )
  on.exit(DBI::dbDisconnect(bench_con, shutdown = TRUE), add = TRUE)

  rducks_enable(bench_con, threads = "single")
  bench_plan <- rducks_execution_plan(
    "arrow_ipc", "multiprocess_parallel",
    future_timeout = 120
  )

  bench_n <- 200000L
  DBI::dbExecute(
    bench_con,
    sprintf(
      "CREATE TABLE bench_input AS SELECT i::BIGINT AS i FROM range(%d) AS t(i)",
      bench_n
    )
  )
  DBI::dbExecute(bench_con, "PRAGMA threads=5")
  DBI::dbExecute(bench_con, "SET external_threads=1")
  parallel_probe <- DBI::dbGetQuery(
    bench_con,
    paste(
      "SELECT count(*) AS rows,",
      "sum(CASE WHEN rducks_thread_is_main(i::UBIGINT) THEN 1 ELSE 0 END) AS main_thread_rows",
      "FROM bench_input"
    )
  )
  stopifnot(parallel_probe$main_thread_rows[[1L]] < parallel_probe$rows[[1L]])

  rducks_set_execution_plan(bench_con, bench_plan, threads = 1L, external_threads = 1L)
  invisible(rducks_register(
    bench_con,
    name = "r_ipc_noop_bench",
    fun = function(x) as.numeric(x) + 1,
    args = BIGINT,
    returns = DOUBLE,
    mode = "vectorized",
    side_effects = FALSE
  ))

  expected <- DBI::dbGetQuery(bench_con, "SELECT sum(i + 1) AS x FROM bench_input")$x[[1L]]
  sql <- "SELECT sum(r_ipc_noop_bench(i)) AS x FROM bench_input"

  time_query <- function(threads) {
    rducks_set_execution_plan(
      bench_con, bench_plan,
      threads = threads,
      external_threads = 1L
    )
    invisible(DBI::dbGetQuery(bench_con, sql)) # warm this execution shape
    elapsed <- system.time(out <- DBI::dbGetQuery(bench_con, sql))[["elapsed"]]
    stopifnot(isTRUE(all.equal(as.numeric(out$x), as.numeric(expected))))
    elapsed
  }

  elapsed <- data.frame(
    mode = c("threads=1", "threads=5"),
    elapsed_seconds = c(time_query(1L), time_query(5L))
  )
  elapsed$speedup_vs_threads_1 <- elapsed$elapsed_seconds[[1L]] / elapsed$elapsed_seconds

  diagnostics <- rducks_explain_udf(bench_con, "r_ipc_noop_bench")[, c(
    "queue_pending_max", "ripc_inflight_max", "ripc_submit_wave_max",
    "ripc_collect_ready_max", "ripc_collect_max_batch"
  )]

  list(
    parallel_probe = parallel_probe,
    elapsed = elapsed,
    diagnostics = diagnostics
  )
})

arrow_ipc_noop_benchmark$parallel_probe
arrow_ipc_noop_benchmark$elapsed
arrow_ipc_noop_benchmark$diagnostics
```

### DuckDB multi-file sleep benchmark

This benchmark uses DuckDB input: four temporary CSV files read by DuckDB's
`read_csv()` table function. The UDF sleeps for 0.50 seconds once per vectorized
DuckDB chunk. The probe again checks that the source runs on a non-main worker
thread before timing the UDF query.

```{r arrow-ipc-csv-sleep-benchmark}
arrow_ipc_sleep_benchmark <- local({
  future::plan(future.mirai::mirai_multisession, workers = 4)

  bench_con <- DBI::dbConnect(
    duckdb::duckdb(config = list(allow_unsigned_extensions = "true")),
    dbdir = ":memory:"
  )
  on.exit(DBI::dbDisconnect(bench_con, shutdown = TRUE), add = TRUE)

  rducks_enable(bench_con, threads = "single")
  bench_plan <- rducks_execution_plan(
    "arrow_ipc", "multiprocess_parallel",
    future_timeout = 120
  )
  rducks_set_execution_plan(bench_con, bench_plan)

  csv_dir <- tempfile("rducks-csv-bench-")
  dir.create(csv_dir)
  csv_files <- file.path(csv_dir, paste0("part", seq_len(4L), ".csv"))
  start <- 0L
  for (file in csv_files) {
    values <- start + seq_len(1024L) - 1L
    writeLines(c("i", as.character(values)), file)
    start <- start + 1024L
  }
  csv_paths <- paste(sprintf("'%s'", csv_files), collapse = ",")
  csv_source <- sprintf(
    "read_csv([%s], types={'i':'BIGINT'}, header=true)",
    csv_paths
  )

  invisible(rducks_register(
    bench_con,
    name = "r_ipc_sleep_half_second_bench",
    fun = function(x) {
      Sys.sleep(0.50)
      as.numeric(x) + 1
    },
    args = BIGINT,
    returns = DOUBLE,
    mode = "vectorized",
    side_effects = TRUE
  ))

  DBI::dbExecute(bench_con, "PRAGMA threads=5")
  DBI::dbExecute(bench_con, "SET external_threads=1")
  parallel_probe <- DBI::dbGetQuery(
    bench_con,
    sprintf(
      paste(
        "SELECT count(*) AS rows,",
        "sum(CASE WHEN rducks_thread_is_main(i::UBIGINT) THEN 1 ELSE 0 END) AS main_thread_rows",
        "FROM %s"
      ),
      csv_source
    )
  )
  stopifnot(parallel_probe$main_thread_rows[[1L]] < parallel_probe$rows[[1L]])

  expected <- 4096L * 4097L / 2
  sql <- sprintf("SELECT sum(r_ipc_sleep_half_second_bench(i)) AS x FROM %s", csv_source)

  time_query <- function(threads) {
    rducks_set_execution_plan(
      bench_con, bench_plan,
      threads = threads,
      external_threads = 1L
    )
    invisible(DBI::dbGetQuery(bench_con, sql)) # warm this execution shape
    elapsed <- system.time(out <- DBI::dbGetQuery(bench_con, sql))[["elapsed"]]
    stopifnot(isTRUE(all.equal(as.numeric(out$x), expected)))
    elapsed
  }

  elapsed <- data.frame(
    mode = c("threads=1", "threads=5"),
    elapsed_seconds = c(time_query(1L), time_query(5L))
  )
  elapsed$speedup_vs_threads_1 <- elapsed$elapsed_seconds[[1L]] / elapsed$elapsed_seconds

  diagnostics <- rducks_explain_udf(bench_con, "r_ipc_sleep_half_second_bench")[, c(
    "queue_pending_max", "ripc_inflight_max", "ripc_submit_wave_max",
    "ripc_collect_ready_max", "ripc_collect_max_batch"
  )]

  list(
    parallel_probe = parallel_probe,
    elapsed = elapsed,
    diagnostics = diagnostics
  )
})

arrow_ipc_sleep_benchmark$parallel_probe
arrow_ipc_sleep_benchmark$elapsed
arrow_ipc_sleep_benchmark$diagnostics
```

These timings are diagnostics, not a portable performance promise. In
particular, cheap UDFs can be slower with `multiprocess_parallel` because Arrow
IPC and Future scheduling overhead dominate much of the work being done.

```{r reset-after-ipc, include=FALSE}
rducks_set_execution_plan(con, rducks_execution_plan("arrow_r", "serial"))
future::plan(future::sequential)
```

## Current scope

Rducks currently builds `rducks.duckdb_extension` at install time, loads it into
DuckDB with `rducks_enable()`, and registers scalar or vectorized R UDFs with
`rducks_register()`.

The input/output type set is `BOOLEAN`, `TINYINT`, `UTINYINT`, `SMALLINT`,
`USMALLINT`, `INTEGER`, `UINTEGER`, `BIGINT`, `UBIGINT`, `FLOAT`, `DOUBLE`,
`VARCHAR`, `BLOB`, `DATE`, `TIME`, `TIMESTAMP`, `HUGEINT`, `UHUGEINT`, `UUID`,
`INTERVAL`, `BIT`, `DECIMAL(width, scale)`, `ENUM(levels)`, and `UNION(...)`.
Composite inputs and outputs are accepted as constructed type objects such as
`TYPE[]`, `TYPE[N]`, `STRUCT(...)`, and `MAP(...)`, recursively over supported
child types.

Enum types are supported by all implemented execution plans. For
`arrow_ipc`, Rducks does not rely on general Arrow dictionary IPC because the
nanoarrow C IPC writer used by this path does not yet encode dictionary arrays.
Instead, enum payloads use the declared Rducks type descriptor as the dictionary
sidecar and serialize the DuckDB enum storage indices as ordinary Arrow integer
arrays. This is an explicit Rducks wire convention for declared `ENUM(...)`
types, not general support for arbitrary Arrow dictionary arrays. A future
same-host payload mode such as `mori` may optimize this path further.

Rducks also provides explicit R value classes for exact or DuckDB-specific
values: `rducks_bigint()`, `rducks_ubigint()`, `rducks_uuid()`,
`rducks_interval()`, `rducks_decimal()`, `rducks_hugeint()`,
`rducks_uhugeint()`, `rducks_bits()`, `rducks_enum()`, and `rducks_union()`.
Constructed DuckDB type objects are formal S7-backed descriptors with structural
validation via `rducks_is_type()`.

## Execution mode semantics

The table below is produced by `rducks_mode_semantics()`.

```{r mode-semantics, echo=FALSE, results="asis"}
mode_semantics <- Rducks::rducks_mode_semantics()
knitr::kable(
  data.frame(
    mode = paste0("`", mode_semantics$mode, "`"),
    status = mode_semantics$status,
    call_granularity = mode_semantics$call_granularity,
    input_shape = mode_semantics$input_shape,
    return_shape = mode_semantics$return_shape,
    length_semantics = mode_semantics$length_semantics,
    threading = mode_semantics$threading,
    copy_semantics = mode_semantics$copy_semantics,
    check.names = FALSE
  ),
  format = "markdown"
)
```

## Type descriptors

```{r type-descriptor-checks}
nested_type <- STRUCT(
  payload = UNION(code = INTEGER, label = ENUM(c("red", "blue"))),
  amount = DECIMAL(10, 2),
  tags = LIST(ENUM(c("red", "blue")))
)

rducks_is_type(nested_type)
rducks_type_kind(nested_type)
rducks_type_sql(nested_type)
rducks_type_child_names(nested_type)
rducks_check_return(UNION(code = INTEGER, label = VARCHAR), rducks_union("label", "ok"))
rducks_check_return(ENUM(c("red", "blue")), rducks_enum("red", c("red", "blue")))
```

### Argument values passed to R functions

<details>
<summary>Expand for argument values passed to R functions</summary>

The table is produced by the exported `rducks_argument_type_mapping()` helper
and reflects the currently implemented nanoarrow scalar marshalling path. With
`null_handling = "default"`, any top-level SQL `NULL` input makes DuckDB return
SQL `NULL` without calling the R function. The `SQL NULL in function` column
below applies when `null_handling = "special"`. It is type-specific: ordinary
R scalar types receive typed `NA` values, while exact/exotic value classes,
binary values, and top-level composite values receive R `NULL`. Within
homogeneous scalar lists/arrays, SQL `NULL` elements are represented as typed
`NA` values where the child type has an R `NA` representation; nested composite
`NULL` values are represented as R `NULL`.

```{r argument-type-mapping, echo=FALSE, results="asis"}
argument_type_mapping <- rbind(
  Rducks::rducks_argument_type_mapping(),
  Rducks::rducks_argument_type_mapping(list(
    INTEGER[],
    BIGINT[3],
    STRUCT(a = UUID, b = DECIMAL(10, 2)),
    MAP(VARCHAR, INTEGER),
    DECIMAL(10, 2),
    ENUM(c("red", "blue")),
    UNION(code = INTEGER, label = VARCHAR)
  ))
)
knitr::kable(
  data.frame(
    argument_type = paste0("`", argument_type_mapping$duckdb_sql, "`"),
    r_type = argument_type_mapping$r_type,
    r_value_passed_to_fun = argument_type_mapping$r_value_passed_to_fun,
    sql_null_in_function = argument_type_mapping$sql_null_in_function,
    copy_semantics = argument_type_mapping$copy_semantics,
    notes = argument_type_mapping$notes,
    check.names = FALSE
  ),
  format = "markdown"
)
```


</details>

### NULL, NA, NaN, Inf, and value-class semantics

<details>
<summary>Expand for NULL, NA, NaN, Inf, and value-class semantics</summary>

The table below is produced by `rducks_value_semantics()`, the exported schema
that Rducks uses to document scalar-mode missing and non-finite value behavior.
Top-level R `NULL` returns map to SQL `NULL`. R `NA` values map to SQL `NULL`
when represented by the declared R type. `NaN` and `Inf` are values only for
`FLOAT` and `DOUBLE`; integer, date, time, timestamp, exact, and exotic return
paths reject them.

```{r value-semantics, echo=FALSE, results="asis"}
value_semantics <- Rducks::rducks_value_semantics(list(
  INTEGER,
  DOUBLE,
  BIGINT,
  UBIGINT,
  HUGEINT,
  UHUGEINT,
  UUID,
  INTERVAL,
  BIT,
  DECIMAL(10, 2),
  ENUM(c("red", "blue")),
  UNION(code = INTEGER, label = VARCHAR),
  STRUCT(amount = DECIMAL(10, 2), id = BIGINT),
  MAP(VARCHAR, INTEGER)
))
knitr::kable(
  data.frame(
    duckdb_type = paste0("`", value_semantics$duckdb_sql, "`"),
    sql_null_special = value_semantics$sql_null_input_special,
    r_na_return = value_semantics$r_na_return,
    r_nan_return = value_semantics$r_nan_return,
    r_inf_return = value_semantics$r_inf_return,
    binary_ops = value_semantics$binary_ops,
    error_semantics = value_semantics$error_semantics,
    check.names = FALSE
  ),
  format = "markdown"
)
```

</details>

## Composite input examples

Homogeneous scalar lists and arrays are passed as atomic R vectors with SQL
`NULL` elements represented as typed `NA` values. Structs are passed as named
lists, and maps are passed as `list(keys = ..., values = ...)`.

```{r composite-inputs}
reg_list_len <- rducks_register(
  con,
  name = "r_list_len",
  fun = function(x) length(x),
  args = INTEGER[],
  returns = INTEGER
)

reg_array_sum <- rducks_register(
  con,
  name = "r_array_sum",
  fun = function(x) sum(x),
  args = INTEGER[3],
  returns = INTEGER
)

reg_struct_sum <- rducks_register(
  con,
  name = "r_struct_sum",
  fun = function(x) x$a + x$b,
  args = STRUCT(a = INTEGER, b = INTEGER),
  returns = INTEGER
)

reg_map_sum <- rducks_register(
  con,
  name = "r_map_sum",
  fun = function(x) sum(x$values),
  args = MAP(VARCHAR, INTEGER),
  returns = INTEGER
)

reg_make_struct <- rducks_register(
  con,
  name = "r_make_struct",
  fun = function(x) list(a = x, b = x + 1L),
  args = INTEGER,
  returns = STRUCT(a = INTEGER, b = INTEGER)
)

reg_make_array <- rducks_register(
  con,
  name = "r_make_array",
  fun = function(x) c(x, x + 1L),
  args = INTEGER,
  returns = INTEGER[]
)

dbGetQuery(con, paste(
  "SELECT",
  "r_list_len([1,2,3]::INTEGER[]) AS list_len,",
  "r_array_sum([1,2,3]::INTEGER[3]) AS array_sum,",
  "r_struct_sum({'a': 20, 'b': 22}::STRUCT(a INTEGER, b INTEGER)) AS struct_sum,",
  "r_map_sum(map(['a','b'], [20,22])) AS map_sum,",
  "(r_make_struct(41::INTEGER)).b AS struct_return,",
  "list_sum(r_make_array(20::INTEGER)) AS array_return"
))
```

## NULL handling

By default, Rducks uses NULL-in/NULL-out handling: if any input is SQL `NULL`,
the R function is not called and the SQL result is `NULL`.

Use `null_handling = "special"` to pass the type-specific missing value shown
in `rducks_argument_type_mapping()` to the R function for SQL `NULL` inputs.
For ordinary scalar types this is usually a typed `NA`; for exact/exotic,
binary, and composite inputs it is R `NULL`.

```{r null-handling}
reg_null_special <- rducks_register(
  con,
  name = "r_null_special",
  fun = function(x) if (is.na(x)) 5L else x,
  args = INTEGER,
  returns = INTEGER,
  null_handling = "special"
)

dbGetQuery(con, "SELECT r_null_special(NULL::INTEGER) AS x")
```

## Exceptions and side effects

Set `exception_handling = "return_null"` to turn errors thrown by the user R
function into SQL `NULL`. Return type-checking and marshalling errors still
abort the query so type bugs are not hidden as NULLs.

```{r exception-handling}
reg_error_null <- rducks_register(
  con,
  name = "r_error_null",
  fun = function(x) stop("boom"),
  args = INTEGER,
  returns = INTEGER,
  exception_handling = "return_null"
)

dbGetQuery(con, "SELECT r_error_null(1::INTEGER) AS x")
```

Set `side_effects = TRUE` for functions with counters, randomness, I/O, or
mutation so DuckDB reruns the function for each row.

```{r side-effects}
counter <- local({
  i <- 0L
  function() {
    i <<- i + 1L
    i
  }
})

reg_counter <- rducks_register(
  con,
  name = "r_counter",
  fun = counter,
  args = character(),
  returns = INTEGER,
  side_effects = TRUE
)

dbGetQuery(con, "SELECT r_counter() AS x FROM range(5)")

set.seed(1)
reg_rng <- rducks_register(
  con,
  name = "r_rng",
  fun = function() runif(1),
  args = character(),
  returns = DOUBLE,
  side_effects = TRUE
)

dbGetQuery(con, "SELECT r_rng() AS x FROM range(3)")
```

## Build notes

The package builds its DuckDB extension during installation using `configure` or
`configure.win`. The extension metadata footer is appended by
`tools/append_extension_metadata.R`.

DuckDB C API headers are refreshed explicitly with:

```sh
Rscript tools/fetch_duckdb_headers.R --ref v1.5.2
```

The DuckDB Arrow C Data path requires DuckDB's unstable C extension API, so the
extension metadata uses `C_STRUCT_UNSTABLE` and must match the bundled DuckDB
header/runtime version. See `docs/BUILD.md` for the extension build and metadata
details.

## DuckDB C extension ABI surface

The extension entrypoint receives the DuckDB database handle from DuckDB's
extension access struct, stores it in a per-database runtime record, and opens an
extension-owned DuckDB connection for native callback support. On extension
reload, Rducks verifies that the active connection has the SQL surface registered
and re-registers it when needed.

The following unstable DuckDB C extension ABI entries are derived by scanning the
Rducks extension C sources and matching `duckdb_*` calls against the versioned
sections of the bundled `duckdb_extension.h`:

```{r used-unstable-duckdb-api, echo=FALSE, results='asis'}
source("tools/used_duckdb_unstable_api.R")
cat(rducks_used_duckdb_unstable_api_markdown())
```

```{r cleanup, include=FALSE}
dbDisconnect(con, shutdown = TRUE)
```
