---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
library(Rducks)
```

# Rducks

Rducks registers R scalar functions as DuckDB SQL functions. It ships as an R
package plus a DuckDB extension. The loaded DuckDB extension registers R functions
on a DuckDB connection and performs the current nanoarrow scalar-mode execution
bridge over DuckDB Arrow C Data.

## How it works

Rducks has two native boundaries: an R package that owns R function lifetime and
registration ergonomics, and a loaded DuckDB extension that owns SQL function
registration, DuckDB chunk access, and current scalar-mode R function execution.

When you call `rducks_enable(con, threads = "single")`, Rducks loads the bundled
`rducks.duckdb_extension` into that DuckDB connection and explicitly sets
`external_threads=1` plus `PRAGMA threads=1`. This is a constraint of R's C API:
Rducks keeps all R API work on the calling R thread. If DuckDB later runs UDF
chunks on worker threads, the extension queues those R execution requests back to
the calling R thread instead of touching the R API from the worker.

When you call `rducks_register()`, Rducks normalizes the declared DuckDB type
objects, checks that scalar-mode marshalling is available, and preserves the R
function. Registration then crosses back through SQL: Rducks calls the extension
function `rducks_register_scalar(...)`, passing the R function reference, type
descriptor tokens, and NULL/exception/side-effect flags. The extension registers
one DuckDB scalar function implementation and stores the per-UDF metadata in
DuckDB `extra_info`. During query execution, that generic DuckDB scalar-function entry point exports
input chunks through the DuckDB Arrow C Data API, calls the private nanoarrow
scalar adapter on the calling R thread, imports the returned Arrow C Data back
into a DuckDB chunk, and attaches the result to the DuckDB output vector.

## Getting started

```{r getting-started}
library(DBI)
library(duckdb)
library(Rducks)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(con, threads = "single")

reg_plus_one <- rducks_register(
  con,
  name = "r_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "scalar"
)

dbGetQuery(con, "SELECT r_plus_one(41.0) AS x")
```

`rducks_register()` returns an `rducks_registration` object that records the
connection, normalized signature, and registration options. You do not need to
keep this object for the UDF to keep working in DuckDB.

The implemented mode is `mode = "scalar"`, which calls the R function once per
DuckDB row. Internally this scalar adapter is nanoarrow-backed over DuckDB Arrow
C Data: DuckDB chunks are exported through Arrow C Data, adapted to scalar R
calls, then imported back to DuckDB. A future vectorized mode should call R once
per DuckDB chunk and will be added only when implemented.

`u32` is passed through R numeric (`double`). `BIGINT`, `UBIGINT`, `HUGEINT`,
and `UHUGEINT` use exact Rducks integer classes backed by canonical decimal
strings.

## Current scope

Rducks currently builds `rducks.duckdb_extension` at install time, loads it into
DuckDB with `rducks_enable()`, and registers scalar-mode R UDFs with
`rducks_register()`. The scalar-mode input/output type set is `BOOLEAN`,
`TINYINT`, `UTINYINT`, `SMALLINT`, `USMALLINT`, `INTEGER`, `UINTEGER`,
`BIGINT`, `UBIGINT`, `FLOAT`, `DOUBLE`, `VARCHAR`, `BLOB`, `DATE`, `TIME`,
`TIMESTAMP`, `HUGEINT`, `UHUGEINT`, `UUID`, `INTERVAL`, `BIT`,
`DECIMAL(width, scale)`, `ENUM(levels)`, and `UNION(...)`. Composite inputs and
outputs are accepted as constructed type objects such as `TYPE[]`, `TYPE[N]`,
`STRUCT(...)`, and `MAP(...)`, recursively over supported child types. The
default `mode = "scalar"` calls R once per DuckDB row. Registration also supports
`null_handling`, `exception_handling`, and `side_effects` controls.

Direct R functions require R API work to happen on the calling R thread. This is
R's thread-affinity rule, not a DuckDB data-race issue. Call
`rducks_enable(con, threads = "single")` or set `external_threads=1` and
`PRAGMA threads=1` before registering R UDFs. Rducks checks this at registration
time.

Rducks also provides explicit R value classes for exact or DuckDB-specific
values: `rducks_bigint()`, `rducks_ubigint()`, `rducks_uuid()`,
`rducks_interval()`, `rducks_decimal()`, `rducks_hugeint()`,
`rducks_uhugeint()`, `rducks_bits()`, `rducks_enum()`, and `rducks_union()`.
These classes preserve exact representation at the scalar-mode R function boundary.
Constructed DuckDB type objects are formal S7-backed
Rducks descriptors with native structural checks via `rducks_is_type()`.
Descriptors are recursive, so lists, arrays, structs, maps, enums, decimals,
and unions can be nested through the constructors rather than quoted type
strings. `rducks_check_argument()` and `rducks_check_return()` can validate
ordinary R values against those descriptors before marshalling.

### Execution mode semantics

The table below is produced by `rducks_mode_semantics()`. Only `scalar` is public
and implemented now.

```{r mode-semantics, echo=FALSE, results="asis"}
mode_semantics <- Rducks::rducks_mode_semantics()
knitr::kable(
  data.frame(
    mode = paste0("`", mode_semantics$mode, "`"),
    status = mode_semantics$status,
    call_granularity = mode_semantics$call_granularity,
    input_shape = mode_semantics$input_shape,
    return_shape = mode_semantics$return_shape,
    length_semantics = mode_semantics$length_semantics,
    threading = mode_semantics$threading,
    copy_semantics = mode_semantics$copy_semantics,
    check.names = FALSE
  ),
  format = "markdown"
)
```

## Type descriptors

```{r type-descriptor-checks}
nested_type <- STRUCT(
  payload = UNION(code = INTEGER, label = ENUM(c("red", "blue"))),
  amount = DECIMAL(10, 2),
  tags = LIST(ENUM(c("red", "blue")))
)

rducks_is_type(nested_type)
rducks_type_kind(nested_type)
rducks_type_sql(nested_type)
rducks_type_child_names(nested_type)
rducks_check_return(UNION(code = INTEGER, label = VARCHAR), rducks_union("label", "ok"))
rducks_check_return(ENUM(c("red", "blue")), rducks_enum("red", c("red", "blue")))
```

### Argument values passed to R functions

<details>
<summary>Expand for argument values passed to R functions</summary>

The table is produced by the exported `rducks_argument_type_mapping()` helper
and reflects the currently implemented nanoarrow scalar marshalling path. With
`null_handling = "default"`, any top-level SQL `NULL` input makes DuckDB return
SQL `NULL` without calling the R function. The `SQL NULL in function` column
below applies when `null_handling = "special"`. It is type-specific: ordinary
R scalar types receive typed `NA` values, while exact/exotic value classes,
binary values, and top-level composite values receive R `NULL`. Within
homogeneous scalar lists/arrays, SQL `NULL` elements are represented as typed
`NA` values where the child type has an R `NA` representation; nested composite
`NULL` values are represented as R `NULL`.

```{r argument-type-mapping, echo=FALSE, results="asis"}
argument_type_mapping <- rbind(
  Rducks::rducks_argument_type_mapping(),
  Rducks::rducks_argument_type_mapping(list(
    INTEGER[],
    BIGINT[3],
    STRUCT(a = UUID, b = DECIMAL(10, 2)),
    MAP(VARCHAR, INTEGER),
    DECIMAL(10, 2),
    ENUM(c("red", "blue")),
    UNION(code = INTEGER, label = VARCHAR)
  ))
)
knitr::kable(
  data.frame(
    argument_type = paste0("`", argument_type_mapping$duckdb_sql, "`"),
    r_type = argument_type_mapping$r_type,
    r_value_passed_to_fun = argument_type_mapping$r_value_passed_to_fun,
    sql_null_in_function = argument_type_mapping$sql_null_in_function,
    copy_semantics = argument_type_mapping$copy_semantics,
    notes = argument_type_mapping$notes,
    check.names = FALSE
  ),
  format = "markdown"
)
```


</details>

### NULL, NA, NaN, Inf, and value-class semantics

<details>
<summary>Expand for NULL, NA, NaN, Inf, and value-class semantics</summary>

The table below is produced by `rducks_value_semantics()`, the exported schema
that Rducks uses to document scalar-mode missing and non-finite value behavior.
Top-level R `NULL` returns map to SQL `NULL`. R `NA` values map to SQL `NULL`
when represented by the declared R type. `NaN` and `Inf` are values only for
`FLOAT` and `DOUBLE`; integer, date, time, timestamp, exact, and exotic return
paths reject them.

```{r value-semantics, echo=FALSE, results="asis"}
value_semantics <- Rducks::rducks_value_semantics(list(
  INTEGER,
  DOUBLE,
  BIGINT,
  UBIGINT,
  HUGEINT,
  UHUGEINT,
  UUID,
  INTERVAL,
  BIT,
  DECIMAL(10, 2),
  ENUM(c("red", "blue")),
  UNION(code = INTEGER, label = VARCHAR),
  STRUCT(amount = DECIMAL(10, 2), id = BIGINT),
  MAP(VARCHAR, INTEGER)
))
knitr::kable(
  data.frame(
    duckdb_type = paste0("`", value_semantics$duckdb_sql, "`"),
    sql_null_special = value_semantics$sql_null_input_special,
    r_na_return = value_semantics$r_na_return,
    r_nan_return = value_semantics$r_nan_return,
    r_inf_return = value_semantics$r_inf_return,
    binary_ops = value_semantics$binary_ops,
    error_semantics = value_semantics$error_semantics,
    check.names = FALSE
  ),
  format = "markdown"
)
```

</details>

## Composite input examples

Homogeneous scalar lists and arrays are passed as atomic R vectors with SQL
`NULL` elements represented as typed `NA` values. Structs are passed as named
lists, and maps are passed as `list(keys = ..., values = ...)`.

```{r composite-inputs}
reg_list_len <- rducks_register(
  con,
  name = "r_list_len",
  fun = function(x) length(x),
  args = INTEGER[],
  returns = INTEGER
)

reg_array_sum <- rducks_register(
  con,
  name = "r_array_sum",
  fun = function(x) sum(x),
  args = INTEGER[3],
  returns = INTEGER
)

reg_struct_sum <- rducks_register(
  con,
  name = "r_struct_sum",
  fun = function(x) x$a + x$b,
  args = STRUCT(a = INTEGER, b = INTEGER),
  returns = INTEGER
)

reg_map_sum <- rducks_register(
  con,
  name = "r_map_sum",
  fun = function(x) sum(x$values),
  args = MAP(VARCHAR, INTEGER),
  returns = INTEGER
)

reg_make_struct <- rducks_register(
  con,
  name = "r_make_struct",
  fun = function(x) list(a = x, b = x + 1L),
  args = INTEGER,
  returns = STRUCT(a = INTEGER, b = INTEGER)
)

reg_make_array <- rducks_register(
  con,
  name = "r_make_array",
  fun = function(x) c(x, x + 1L),
  args = INTEGER,
  returns = INTEGER[]
)

dbGetQuery(con, paste(
  "SELECT",
  "r_list_len([1,2,3]::INTEGER[]) AS list_len,",
  "r_array_sum([1,2,3]::INTEGER[3]) AS array_sum,",
  "r_struct_sum({'a': 20, 'b': 22}::STRUCT(a INTEGER, b INTEGER)) AS struct_sum,",
  "r_map_sum(map(['a','b'], [20,22])) AS map_sum,",
  "(r_make_struct(41::INTEGER)).b AS struct_return,",
  "list_sum(r_make_array(20::INTEGER)) AS array_return"
))
```

## NULL handling

By default, Rducks uses NULL-in/NULL-out handling: if any input is SQL `NULL`,
the R function is not called and the SQL result is `NULL`.

Use `null_handling = "special"` to pass the type-specific missing value shown
in `rducks_argument_type_mapping()` to the R function for SQL `NULL` inputs.
For ordinary scalar types this is usually a typed `NA`; for exact/exotic,
binary, and composite inputs it is R `NULL`.

```{r null-handling}
reg_null_special <- rducks_register(
  con,
  name = "r_null_special",
  fun = function(x) if (is.na(x)) 5L else x,
  args = INTEGER,
  returns = INTEGER,
  null_handling = "special"
)

dbGetQuery(con, "SELECT r_null_special(NULL::INTEGER) AS x")
```

## Exceptions and side effects

Set `exception_handling = "return_null"` to turn R errors into SQL
`NULL`.

```{r exception-handling}
reg_error_null <- rducks_register(
  con,
  name = "r_error_null",
  fun = function(x) stop("boom"),
  args = INTEGER,
  returns = INTEGER,
  exception_handling = "return_null"
)

dbGetQuery(con, "SELECT r_error_null(1::INTEGER) AS x")
```

Set `side_effects = TRUE` for functions with counters, randomness, I/O, or
mutation so DuckDB reruns the function for each row.

```{r side-effects}
counter <- local({
  i <- 0L
  function() {
    i <<- i + 1L
    i
  }
})

reg_counter <- rducks_register(
  con,
  name = "r_counter",
  fun = counter,
  args = character(),
  returns = INTEGER,
  side_effects = TRUE
)

dbGetQuery(con, "SELECT r_counter() AS x FROM range(5)")

set.seed(1)
reg_rng <- rducks_register(
  con,
  name = "r_rng",
  fun = function() runif(1),
  args = character(),
  returns = DOUBLE,
  side_effects = TRUE
)

dbGetQuery(con, "SELECT r_rng() AS x FROM range(3)")
```

## Build notes

The package builds its DuckDB extension during installation using `configure` or
`configure.win`. The extension metadata footer is appended by
`tools/append_extension_metadata.R`.

DuckDB C API headers are refreshed explicitly with:

```sh
Rscript tools/fetch_duckdb_headers.R --ref v1.5.2
```

The DuckDB Arrow C Data path requires DuckDB's unstable C extension API, so the
extension metadata uses `C_STRUCT_UNSTABLE` and must match the bundled DuckDB header/runtime
version. See `docs/BUILD.md` for the extension build and metadata details.

```{r cleanup, include=FALSE}
dbDisconnect(con, shutdown = TRUE)
```
