<!-- README.md is generated from README.Rmd. Please edit that file and run `make rdm`. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Rducks

Rducks is an R package plus DuckDB extension for registering R functions as
DuckDB scalar UDFs. DuckDB vectors are materialized directly to R values on the
recorded R thread, with no Arrow/nanoarrow intermediary.

## Quick example

```{r example}
library(Rducks)

db <- duckdb::dbConnect(duckdb::duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(db, threads = "single")

rducks_register_scalar_udf(
  db, "plus_one", function(x) x + 1L,
  args = list(INTEGER), returns = INTEGER
)

DBI::dbGetQuery(db, "SELECT plus_one(41::INTEGER) AS x")

rducks_release(db)
DBI::dbDisconnect(db, shutdown = TRUE)
```

## Status

The supported execution path is direct in-process marshalling. The experimental
Quack wire codec (`src/quack_core.c`, `src/quack_codec.c`) is covered by
tinytests as the replacement wire-format foundation, but the worker-process IPC
transport is intentionally not advertised until its native DuckDB-vector adapter
is implemented.
