---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
library(Rducks)
```

# Rducks

[![R-CMD-check](https://github.com/sounkou-bioinfo/Rducks/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rducks/actions/workflows/R-CMD-check.yaml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Rducks)](https://sounkou-bioinfo.r-universe.dev/Rducks)

Rducks registers R functions as DuckDB SQL functions using a DuckDB C
extension, including a small set of unstable DuckDB C extension API functions.
The extension records the DuckDB database instance handle at initialization and
keeps an extension-owned connection associated with that runtime.

The user-facing API separates three concepts:

- DuckDB function kind: scalar UDF, aggregate function, or table function.
- Rducks evaluation mode for DuckDB scalar UDFs: one R call per row
  (`mode = "scalar"`) or one R call per DuckDB chunk (`mode = "vectorized"`).
- Rducks execution plan: the marshalling and concurrency implementation used by
  DuckDB scalar UDFs.
- R-side query consumption: `rducks_query_stream()` exposes explicit
  connection-bound query batches; it is not a SQL UDF registration surface.

R-backed DuckDB scalar UDF inputs and outputs move through explicit execution
plans:

- `arrow_r`: reference path using DuckDB Arrow C Data plus nanoarrow/R.
- `arrow_c`: native extension path for supported scalar-UDF evaluation modes
  with direct DuckDB-vector materialization.
- `arrow_ipc`: process-isolated R execution using native NNG plus owned
  Arrow IPC request/result bytes. By default Rducks launches worker loops with
  mirai daemons and Rducks-generated NNG endpoint URLs; `ipc_transport` selects
  the generated transport. If `ipc_endpoints` is supplied, Rducks connects to
  worker URLs that the caller starts and stops. Large same-host read-only R
  globals can optionally be sent as mori shared-memory references with
  `ipc_globals_share = "mori"`.

Aggregate and table functions use separate APIs and are not scalar-UDF
execution-plan variants. `rducks_register_aggregate()` stores only native
raw-byte aggregate state and calls R for serialized update/finalize phases on
the recorded R thread. `rducks_register_table()` covers finite scans whose
positional SQL argument count is inferred from the R function formals and whose
input types are registered as DuckDB `ANY`; the output schema is inferred during
DuckDB bind and the result is imported through nanoarrow Arrow C Data before
scanning. For R callers that want query results incrementally,
`rducks_query_stream()` provides a connection-bound query-batch API with
explicit `next_batch()` and `close()` methods. Query streams fetch DuckDB native
chunks and import them through Arrow C Data/nanoarrow without requiring the
heavy `arrow` R package. The current R-facing `next_batch()` method materializes
base-R batches, but the stream engine boundary is DuckDB chunks and Arrow C
Data rather than DBI result pagination.

## Quick start

```{r getting-started}
library(DBI)
library(duckdb)
library(Rducks)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(con, threads = "single")

reg_hello <- rducks_register_scalar_udf(
  con,
  name = "r_hello",
  fun = function() "hello from R",
  args = NULL,
  returns = VARCHAR,
  mode = "scalar"
)

dbGetQuery(con, "SELECT r_hello() AS message")

reg_plus_one <- rducks_register_scalar_udf(
  con,
  name = "r_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "scalar"
)

dbGetQuery(con, "SELECT r_plus_one(41.0) AS x")
```

Use `args = NULL` for a zero-argument scalar UDF. `null_handling` only affects
SQL NULL inputs, so the default is appropriate for `r_hello()`.

`rducks_register_scalar_udf()` returns an `rducks_scalar_udf_registration` object that records the
connection, normalized signature, and registration options. You do not need to
keep this object for the scalar UDF to keep working in DuckDB. Registering the same SQL
name/signature again replaces the callable implementation in the shared DuckDB
database catalog.

### Release and disconnect

`rducks_release(con)` detaches connection-local Rducks state such as the current
default plan and R-side runtime anchor. It is non-destructive for the DuckDB
catalog: it does not drop catalog functions and does not release closures still
owned by native catalog metadata. For `arrow_ipc + multiprocess_parallel`, the
last Rducks attachment to a runtime also closes native client pools for
Rducks-launched local workers and stops those local mirai/NNG workers. If
`ipc_endpoints` was supplied, those URLs name user-owned worker processes;
Rducks does not send stop requests to them during release. Call
`rducks_release(con)` before `DBI::dbDisconnect(con)` when you want deterministic
R-side cleanup. Rducks does not provide `rducks_unregister()`; registered
catalog functions and their preserved R closures are intentionally retained for
the database/runtime lifetime or until the same SQL name/signature is replaced.

```{r release-example}
release_con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(release_con, threads = "single")
rducks_release(release_con)
dbDisconnect(release_con, shutdown = TRUE)
```

### Arrow conversion setting

`rducks_enable()` sets DuckDB's `arrow_lossless_conversion=true` setting on the
user connection, and the Rducks extension also sets it on its own internal
DuckDB connection. Rducks uses DuckDB's Arrow C Data conversion APIs for several
execution plans, and this setting tells DuckDB to preserve DuckDB-specific type
identity in Arrow metadata when possible instead of exporting only a more generic
Arrow representation. This matters for types such as `UUID`, `BIT`,
`HUGEINT`/`UHUGEINT`, `TIME WITH TIME ZONE`, and JSON aliases. It does not
change stored DuckDB values or SQL evaluation; it only affects DuckDB-to-Arrow
conversion used at the Rducks boundary.

## What you can register

Rducks installs a package-managed DuckDB extension and loads it into DuckDB
with `rducks_enable()`. The type descriptors below are used by
`rducks_register_scalar_udf()` for DuckDB scalar UDF arguments/returns and by
`rducks_register_aggregate()` for aggregate inputs and returns. Table functions
registered with `rducks_register_table()` infer their output schema dynamically
from the returned data frame or named list instead of taking declared type
descriptors.

Supported input/output descriptors are `BOOLEAN`, `TINYINT`, `UTINYINT`,
`SMALLINT`, `USMALLINT`, `INTEGER`, `UINTEGER`, `BIGINT`, `UBIGINT`, `FLOAT`,
`DOUBLE`, `VARCHAR`, `BLOB`, `DATE`, `TIME`, `TIMESTAMP`, `HUGEINT`,
`UHUGEINT`, `UUID`, `INTERVAL`, `BIT`, `DECIMAL(width, scale)`,
`ENUM(levels)`, and `UNION(...)`.
Composite inputs and outputs are accepted as constructed type descriptors such
as `TYPE[]`, `TYPE[N]`, `STRUCT(...)`, and `MAP(...)`, recursively over
supported child types.

Declared `ENUM(...)` descriptors are supported by all implemented execution
plans. On the native `arrow_ipc` NNG path, Rducks does not depend on Arrow IPC
dictionary transport for enums: declared enum levels are registration metadata,
and the IPC payload carries the underlying DuckDB enum index storage as Arrow
integer buffers.

Rducks also provides explicit R value classes for exact or DuckDB-specific
values: `rducks_bigint()`, `rducks_ubigint()`, `rducks_uuid()`,
`rducks_interval()`, `rducks_decimal()`, `rducks_hugeint()`,
`rducks_uhugeint()`, `rducks_bits()`, `rducks_enum()`, and `rducks_union()`.
Constructed DuckDB type descriptors use the formal S7 `rducks_type` class and
structural validation via `rducks_is_type()`.

## Type descriptors

```{r type-descriptor-checks}
nested_type <- STRUCT(
  payload = UNION(code = INTEGER, label = ENUM(c("red", "blue"))),
  amount = DECIMAL(10, 2),
  tags = LIST(ENUM(c("red", "blue")))
)

rducks_is_type(nested_type)
rducks_type_kind(nested_type)
rducks_type_sql(nested_type)
rducks_type_child_names(nested_type)
rducks_check_return(UNION(code = INTEGER, label = VARCHAR), rducks_union("label", "ok"))
rducks_check_return(ENUM(c("red", "blue")), rducks_enum("red", c("red", "blue")))
```

## DuckDB scalar UDFs (`rducks_register_scalar_udf()`)

DuckDB scalar UDFs return one SQL value per logical input row. Rducks can
evaluate the backing R function row-wise (`mode = "scalar"`) or chunk-wise
(`mode = "vectorized"`). Vectorized mode calls the R function once per DuckDB
chunk with one R vector/list-column per declared argument.

```{r vectorized-mode}
reg_vec_plus_one <- rducks_register_scalar_udf(
  con,
  name = "r_vec_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "vectorized",
  side_effects = TRUE
)

dbGetQuery(con, "SELECT sum(r_vec_plus_one(i::DOUBLE)) AS x FROM range(5) AS t(i)")
```

In vectorized evaluation mode, `null_handling = "default"` evaluates only rows
with no top-level SQL NULL inputs and scatters SQL NULLs back into the result.
`null_handling = "special"` passes all rows using the same NA/NULL shapes as
row-wise scalar evaluation. The return length must match the number of evaluated rows.
Vectorized mode currently requires at least one declared argument.

### Scalar-UDF evaluation-mode benchmark

A tiny benchmark with `bench` can show the row-wise versus chunk-wise R call
shape for a DuckDB scalar UDF. The result is illustrative rather than a
performance guarantee.

```{r vectorized-benchmark}
bench::mark(
  scalar = dbGetQuery(con, "SELECT sum(r_plus_one(i::DOUBLE)) AS x FROM range(10000) AS t(i)"),
  vectorized = dbGetQuery(con, "SELECT sum(r_vec_plus_one(i::DOUBLE)) AS x FROM range(10000) AS t(i)"),
  iterations = 3,
  check = FALSE
)[, c("expression", "median", "itr/sec", "mem_alloc")]
```

## Scalar-UDF evaluation-mode semantics

The table below is produced by `rducks_mode_semantics()` and applies to DuckDB
scalar UDFs registered with `rducks_register_scalar_udf()`.

```{r mode-semantics, echo=FALSE, results="asis"}
knitr::kable(rducks_mode_semantics(), format = "markdown")
```

## Scalar-UDF argument values passed to R functions

<details>
<summary>Expand for scalar-UDF argument values passed to R functions</summary>

The table is produced by the exported `rducks_argument_type_mapping()` helper
and reflects the currently implemented DuckDB scalar-UDF marshalling path. With
`null_handling = "default"`, any top-level SQL `NULL` input makes DuckDB return
SQL `NULL` without calling the R function. The `special_null_argument` column
below applies when `null_handling = "special"`. It is type-specific: ordinary
R scalar types receive typed `NA` values, while exact Rducks value classes,
binary values, and top-level composite values receive R `NULL`. Within
homogeneous scalar lists/arrays, SQL `NULL` elements are represented as typed
`NA` values where the child type has an R `NA` representation; nested composite
`NULL` values are represented as R `NULL`.

```{r argument-type-mapping, echo=FALSE, results="asis"}
knitr::kable(rducks_argument_type_mapping(list(
  INTEGER,
  DOUBLE,
  INTEGER[],
  BIGINT[3],
  STRUCT(a = UUID, b = DECIMAL(10, 2)),
  MAP(VARCHAR, INTEGER),
  ENUM(c("red", "blue")),
  UNION(code = INTEGER, label = VARCHAR)
)), format = "markdown")
```


</details>

## Scalar-UDF NULL, NA, NaN, Inf, and value-class semantics

<details>
<summary>Expand for scalar-UDF NULL, NA, NaN, Inf, and value-class semantics</summary>

The table below is produced by `rducks_value_semantics()`, the exported schema
that Rducks uses to document DuckDB scalar-UDF missing and non-finite value
behavior.
Top-level R `NULL` returns map to SQL `NULL`. R `NA` values map to SQL `NULL`
when represented by the declared R type. `NaN` and `Inf` are values only for
`FLOAT` and `DOUBLE`; integer, date, time, timestamp, and exact Rducks value
classes reject them.

```{r value-semantics, echo=FALSE, results="asis"}
knitr::kable(rducks_value_semantics(list(
  INTEGER,
  DOUBLE,
  BIGINT,
  UUID,
  DECIMAL(10, 2),
  ENUM(c("red", "blue")),
  UNION(code = INTEGER, label = VARCHAR),
  MAP(VARCHAR, INTEGER)
)), format = "markdown")
```

</details>

## Scalar-UDF composite input examples

Homogeneous scalar lists and arrays are passed as atomic R vectors with SQL
`NULL` elements represented as typed `NA` values. Structs are passed as named
lists, and maps are passed as `list(keys = ..., values = ...)`. A `data.frame`
return is not a table-valued UDF here; the scalar UDF type equivalent is
`STRUCT(...)`. In vectorized evaluation mode, a `STRUCT(...)` return may be
supplied as a `data.frame` with columns matching the struct fields, yielding one
struct-valued SQL result per row.

```{r composite-inputs}
reg_list_len <- rducks_register_scalar_udf(
  con,
  name = "r_list_len",
  fun = function(x) length(x),
  args = INTEGER[],
  returns = INTEGER
)

reg_array_sum <- rducks_register_scalar_udf(
  con,
  name = "r_array_sum",
  fun = function(x) sum(x),
  args = INTEGER[3],
  returns = INTEGER
)

reg_struct_sum <- rducks_register_scalar_udf(
  con,
  name = "r_struct_sum",
  fun = function(x) x$a + x$b,
  args = STRUCT(a = INTEGER, b = INTEGER),
  returns = INTEGER
)

reg_map_sum <- rducks_register_scalar_udf(
  con,
  name = "r_map_sum",
  fun = function(x) sum(x$values),
  args = MAP(VARCHAR, INTEGER),
  returns = INTEGER
)

reg_make_struct <- rducks_register_scalar_udf(
  con,
  name = "r_make_struct",
  fun = function(x) list(a = x, b = x + 1L),
  args = INTEGER,
  returns = STRUCT(a = INTEGER, b = INTEGER)
)

reg_make_array <- rducks_register_scalar_udf(
  con,
  name = "r_make_array",
  fun = function(x) c(x, x + 1L),
  args = INTEGER,
  returns = INTEGER[]
)

reg_make_map <- rducks_register_scalar_udf(
  con,
  name = "r_make_map",
  fun = function(x) list(keys = c("a", "b"), values = c(x, x + 1L)),
  args = INTEGER,
  returns = MAP(VARCHAR, INTEGER)
)

dbGetQuery(con, "
  SELECT
    r_list_len([1,2,3]::INTEGER[]) AS list_len,
    r_array_sum([1,2,3]::INTEGER[3]) AS array_sum,
    r_struct_sum({'a': 20, 'b': 22}::STRUCT(a INTEGER, b INTEGER)) AS struct_sum,
    r_map_sum(map(['a','b'], [20,22])) AS map_sum,
    (r_make_struct(41::INTEGER)).b AS struct_return,
    list_sum(r_make_array(20::INTEGER)) AS array_return,
    map_extract_value(r_make_map(20::INTEGER), 'b') AS map_return
")
```

## Scalar-UDF NULL handling

For DuckDB scalar UDFs, Rducks uses NULL-in/NULL-out handling by default:
if any input is SQL `NULL`, the R function is not called and the SQL result is
`NULL`.

Use `null_handling = "special"` to pass the type-specific missing value shown
in `rducks_argument_type_mapping()` to the R function for SQL `NULL` inputs.
For ordinary scalar types this is usually a typed `NA`; for exact/exotic,
binary, and composite inputs it is R `NULL`.

```{r null-handling}
reg_null_special <- rducks_register_scalar_udf(
  con,
  name = "r_null_special",
  fun = function(x) if (is.na(x)) 5L else x,
  args = INTEGER,
  returns = INTEGER,
  null_handling = "special"
)

dbGetQuery(con, "SELECT r_null_special(NULL::INTEGER) AS x")
```

## Scalar-UDF exceptions and side effects

For DuckDB scalar UDFs, set `exception_handling = "return_null"` to turn
errors thrown by the user R function into SQL `NULL`. Return type-checking and
marshalling errors still abort the query so type bugs are not hidden as NULLs.

```{r exception-handling}
reg_error_null <- rducks_register_scalar_udf(
  con,
  name = "r_error_null",
  fun = function(x) stop("boom"),
  args = INTEGER,
  returns = INTEGER,
  exception_handling = "return_null"
)

dbGetQuery(con, "SELECT r_error_null(1::INTEGER) AS x")
```

For DuckDB scalar UDFs, set `side_effects = TRUE` for functions with
counters, randomness, I/O, or mutation so DuckDB reruns the function for each
row.

```{r side-effects}
counter <- local({
  i <- 0L
  function() {
    i <<- i + 1L
    i
  }
})

reg_counter <- rducks_register_scalar_udf(
  con,
  name = "r_counter",
  fun = counter,
  args = character(),
  returns = INTEGER,
  side_effects = TRUE
)

dbGetQuery(con, "SELECT r_counter() AS x FROM range(5)")

set.seed(1)
reg_rng <- rducks_register_scalar_udf(
  con,
  name = "r_rng",
  fun = function() runif(1),
  args = character(),
  returns = DOUBLE,
  side_effects = TRUE
)

dbGetQuery(con, "SELECT r_rng() AS x FROM range(3)")
```

## Aggregate functions

`rducks_register_aggregate()` registers serialized R-backed aggregates. The
state stored in DuckDB is a native copy of the raw vector returned by
`update(state, ...)`; R object pointers are not stored in aggregate state.
`finalize(state)` receives the final raw state or `NULL` for a group with no
non-NULL inputs and returns a scalar compatible with the declared return type.
Aggregate functions expose a single aggregate execution path: serialized R
callbacks on the recorded R thread. They do not have row-wise/vectorized
scalar-UDF evaluation modes and are not controlled by scalar-UDF execution
plans. Registration requires
`rducks_enable(con, threads = "single")`, and execution rejects attempts to call
R from DuckDB worker threads.

```{r aggregate-mode}
pack_state <- function(x) serialize(x, NULL, version = 2)
unpack_state <- function(state) {
  if (is.null(state)) list(sum = 0L, n = 0L) else unserialize(state)
}

reg_r_sum <- rducks_register_aggregate(
  con,
  name = "r_sum_i32",
  update = function(state, x) {
    s <- unpack_state(state)
    s$sum <- as.integer(s$sum + x)
    s$n <- as.integer(s$n + 1L)
    pack_state(s)
  },
  finalize = function(state) {
    s <- unpack_state(state)
    if (identical(s$n, 0L)) NA_integer_ else as.integer(s$sum)
  },
  args = INTEGER,
  returns = INTEGER
)

dbGetQuery(
  con,
  "SELECT r_sum_i32(i) AS s FROM (VALUES (1::INTEGER), (2::INTEGER), (NULL::INTEGER)) t(i)"
)
```

## Table functions

`rducks_register_table()` registers finite R-backed table functions. The number
of positional SQL arguments comes from `formals(fun)`, each input is registered
as DuckDB `ANY`, and the output schema is inferred from the data frame or named
list returned during DuckDB bind. Table functions expose a single table-function
execution path: R callbacks on the recorded R thread. They do not have
row-wise/vectorized scalar-UDF evaluation modes and are not controlled by
scalar-UDF execution plans; register and
execute them with `rducks_enable(con, threads = "single")`.

```{r table-function}
reg_rows <- rducks_register_table(
  con,
  name = "r_rows",
  fun = function(n, prefix) {
    i <- seq_len(as.integer(n))
    data.frame(i = i, label = paste0(prefix, i))
  },
  chunk_size = 2L
)

dbGetQuery(con, "SELECT * FROM r_rows(3, 'row-') ORDER BY i")
```

## Streaming query batches

Use `rducks_query_stream()` when an R caller wants explicit batch-by-batch query
consumption rather than one eager `DBI::dbGetQuery()` result. The query runs
through the Rducks extension's DuckDB streaming result handle; fetched DuckDB
chunks are exported through DuckDB Arrow C Data and materialized with the same
Rducks/nanoarrow conversion helpers used by scalar-UDF marshalling. This path
does not page over an already materialized DBI result and does not require the
heavy `arrow` R package. Because execution uses the extension-owned DuckDB
connection, database-scoped objects are visible, while caller-connection
temporary tables/views are not part of the stream query scope. Delivery into R
is intentionally on the recorded R thread because the current materializer
creates R objects; Rducks does not call R/nanoarrow materializers from arbitrary
DuckDB worker threads. The returned stream is connection-bound for lifecycle and
has `next_batch()`, `close()`, and `is_closed()` methods. The current
`next_batch()` materializer returns a base R data-frame batch or `NULL` at
end-of-stream. Each non-empty batch carries the stream's nanoarrow schema in the
`"rducks_nanoarrow_schema"` attribute, and `rducks_release(con)` closes streams
attached to that connection.

```{r streaming-query}
stream <- rducks_query_stream(
  con,
  "SELECT i::INTEGER AS i FROM range(1, 6) t(i)",
  batch_size = 2L
)
stream$next_batch()
stream$next_batch()
stream$close()
```

## Execution plans

Execution plans choose the default marshalling implementation and concurrency
contract for future DuckDB scalar UDF registrations through
`rducks_register_scalar_udf()`. They do not control aggregate functions, table
functions, or query streams. Registration remains semantic: name, function,
Rducks evaluation mode, types, NULL handling, exception handling, and
side-effect flag. The plan active at `rducks_register_scalar_udf()` freezes the
scalar UDF's native evaluator/marshalling metadata; later plan changes do not
retarget already-registered scalar UDFs.

| Plan | Scalar | Vectorized | Notes |
| --- | --- | --- | --- |
| `arrow_r + serial` | implemented | implemented | reference implementation |
| `arrow_r + inproc_concurrent` | implemented | implemented | queued same-process callbacks; R API work stays on the recorded main R thread |
| `arrow_c + serial` | implemented | implemented | direct native evaluator tokens `RC`/`RCV` |
| `arrow_c + inproc_concurrent` | implemented | implemented | queued same-process callbacks with direct `arrow_c` marshalling |
| `arrow_ipc + multiprocess_parallel` | implemented | implemented | native NNG plus owned Arrow IPC bytes; mirai-launched local workers by default; `ipc_transport` generates `abstract` (Linux abstract IPC), `ipc` (NNG IPC), `unix` (POSIX Unix-domain alias), `tcp`, or `ws` endpoints; optional explicit `ipc_endpoints`; strict plan |

`arrow_r + serial` is the semantic reference. Other implemented plans are tested
against it and use exactly their selected marshalling path. Use
`rducks_explain_udf()` to inspect the plan and native counters for a registered
scalar UDF.

```{r explain-udf}
rducks_explain_udf(con, "r_vec_plus_one")[, c(
  "name", "mode", "plan_id", "native_marshalling",
  "evaluator", "arrow_r_chunks", "arrow_c_chunks", "arrow_ipc_chunks"
)]
```

## In-process queued execution

`rducks_enable_inproc()` lets DuckDB worker threads enqueue same-process R calls
for scalar UDFs while the recorded main R thread drains the queue. This is for
safety and liveness, not automatic speed: R API work still runs on the main
thread.

```{r inproc-queue}
rducks_register_scalar_udf(
  con,
  name = "r_inproc_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE
)

rducks_enable_inproc(con, threads = 4, external_threads = 1)
dbExecute(con, "CREATE TEMP TABLE inproc_input AS SELECT i::DOUBLE AS x FROM range(200000) AS t(i)")
dbGetQuery(con, "SELECT sum(r_inproc_plus_one(x)) AS total FROM inproc_input")
rducks_inproc_stats(con)[, c("submitted", "executed", "pending_max", "main_drain_batches")]
rducks_disable_inproc(con, threads = 1)
```

## Multiprocess Arrow IPC execution

`arrow_ipc + multiprocess_parallel` is the native NNG/Arrow IPC path: DuckDB
chunks become owned Arrow IPC bytes, native NNG sends them to R worker loops, and
Arrow IPC result bytes are imported back into DuckDB output vectors. Local
workers are mirai daemons running Rducks worker loops. If `ipc_endpoints` is
supplied, Rducks connects to worker URLs that the caller starts and stops.

Operationally:

- `ipc_workers` is the number of persistent R worker processes.
- `ipc_transport` chooses local mirai worker endpoints: `abstract`, `ipc`,
  `unix`, `tcp`, or `ws`.
- `ipc_timeout` is used for worker registration/control requests and for native
  NNG send/receive on each scalar UDF's client pool.
- Worker providers are reused by runtime, worker count, max-pending limit, and
  endpoint/transport choice. Registering another scalar UDF with the same IPC
  provider key reuses the existing workers; changing that key starts another
  provider. Changing the default plan does not stop old providers.
- `rducks_release(con)` stops Rducks-launched local providers when the last
  Rducks anchor for the DuckDB runtime is released. For caller-supplied
  `ipc_endpoints`, release closes Rducks bookkeeping but does not send stop
  requests to the processes listening at those URLs. `rducks_nng_quiesce()` is
  lower-level: it only closes native client pools for local IPC providers.

### Large IPC globals with mori

Automatic global discovery for parallel workers is subtle. Rducks uses
`globals::globalsOf(..., method = "dfs", recursive = TRUE)` when available,
following the futureverse fix around `globals::findGlobals(..., method =
"dfs")`; see Henrik Bengtsson's discussion of why conservative
`codetools`-style discovery can miss globals needed by worker processes:
<https://www.jottr.org/2025/06/23/future-got-better-at-finding-global-variables/>.
For production UDFs with large dependencies, prefer explicit `ipc_globals`.
When the workers are on the same host, `ipc_globals_share = "mori"` applies
`mori::share()` to each selected global before worker registration so large
atomic vectors, lists, and data frames travel as compact shared-memory
references rather than full serialized copies. Rducks keeps those shared objects
anchored for the registered scalar UDF lifetime.

```{r ipc-mori-globals, eval=FALSE}
large_lookup <- rnorm(1e6)

mori_ipc_plan <- rducks_execution_plan(
  "arrow_ipc", "multiprocess_parallel",
  ipc_workers = 2L,
  ipc_globals = "large_lookup",
  ipc_globals_share = "mori"
)
rducks_set_execution_plan(con, mori_ipc_plan, threads = 1, external_threads = 1)

rducks_register_scalar_udf(
  con,
  name = "r_with_large_lookup",
  fun = function(x) x + large_lookup[[1L]],
  args = DOUBLE,
  returns = DOUBLE,
  mode = "vectorized",
  side_effects = TRUE
)
```

This mori path is only for sharing R globals. Built-in IPC backends do not yet
support SQL chunk shared-memory handles. A future shared-memory data plane for
DuckDB chunks should use Rducks-owned raw byte buffers and NNG handle messages,
not per-chunk ALTREP objects or R C API calls from DuckDB worker threads. See
`docs/SHARED_MEMORY_IPC.md` and the local diagnostic
`tools/benchmark_ipc_data_plane.R` before adding any user-facing data-plane
option.

The example below registers the same vectorized R function three ways. Keep
`threads = 1` while registering; raise DuckDB `threads` for the IPC query phase
so chunks can fan out to the IPC workers. `ripc_inflight_max` reports observed
native IPC concurrency and is scheduler-dependent.

```{r arrow-ipc-multiprocess}
chunk_plus_one <- function(x) {
  Sys.sleep(0.05) # visible chunk work for the comparison, not required by Rducks
  x + 1L
}

csv_dir <- file.path(tempdir(), paste0("rducks-ipc-csv-", Sys.getpid()))
dir.create(csv_dir, showWarnings = FALSE)
rows_per_file <- 4096L
parts <- 8L
for (part in seq_len(parts)) {
  values <- seq.int((part - 1L) * rows_per_file, part * rows_per_file - 1L)
  writeLines(c("i", as.character(values)), file.path(csv_dir, sprintf("part-%02d.csv", part)))
}
csv_glob <- file.path(csv_dir, "part-*.csv")

serial_plan <- rducks_execution_plan("arrow_r", "serial")
inproc_plan <- rducks_execution_plan("arrow_r", "inproc_concurrent")
ipc_workers <- 2L
ipc_plan <- rducks_execution_plan(
  "arrow_ipc", "multiprocess_parallel",
  ipc_transport = "tcp",
  ipc_workers = ipc_workers,
  ipc_timeout = 30
)

register_for_plan <- function(name, plan) {
  rducks_set_execution_plan(con, plan, threads = 1, external_threads = 1)
  invisible(rducks_register_scalar_udf(
    con,
    name = name,
    fun = chunk_plus_one,
    args = INTEGER,
    returns = INTEGER,
    mode = "vectorized",
    side_effects = TRUE
  ))
}

register_for_plan("r_cmp_serial", serial_plan)
register_for_plan("r_cmp_inproc", inproc_plan)
register_for_plan("r_cmp_ipc", ipc_plan)

run_comparison <- function(label, name, plan, threads) {
  rducks_set_execution_plan(con, plan, threads = threads, external_threads = 1)
  elapsed <- system.time({
    result <- dbGetQuery(con, sprintf(
      "SELECT sum(%s(i::INTEGER)) AS total FROM read_csv_auto(%s, header = true)",
      DBI::dbQuoteIdentifier(con, name),
      DBI::dbQuoteString(con, csv_glob)
    ))
  })[["elapsed"]]
  info <- rducks_explain_udf(con, name)
  data.frame(
    plan = label,
    threads = threads,
    total = result$total[[1]],
    elapsed_sec = round(unname(elapsed), 3),
    evaluator = info$evaluator[[1]],
    arrow_r_chunks = info$arrow_r_chunks[[1]],
    arrow_ipc_chunks = info$arrow_ipc_chunks[[1]],
    ripc_inflight_max = info$ripc_inflight_max[[1]],
    stringsAsFactors = FALSE
  )
}

comparison <- rbind(
  run_comparison("sequential arrow_r", "r_cmp_serial", serial_plan, threads = 1),
  run_comparison("in-process queue", "r_cmp_inproc", inproc_plan, threads = 1),
  run_comparison("2-process Arrow IPC", "r_cmp_ipc", ipc_plan, threads = ipc_workers)
)
comparison

unlink(csv_dir, recursive = TRUE, force = TRUE)
rducks_set_execution_plan(con, serial_plan, threads = 1, external_threads = 1)
```

## Build notes

The package builds its DuckDB extension during installation using `configure` or
`configure.win`. The extension metadata footer is appended by
`tools/append_extension_metadata.R`.

DuckDB C API headers are refreshed explicitly with:

```sh
Rscript tools/fetch_duckdb_headers.R --ref v1.5.2
```

The DuckDB Arrow C Data path requires DuckDB's unstable C extension API, so the
extension metadata uses `C_STRUCT_UNSTABLE` and must match the bundled DuckDB
header/runtime version. See `docs/BUILD.md` for the extension build and metadata
details.

## DuckDB C extension ABI surface

The extension entrypoint receives the DuckDB database handle from DuckDB's
extension access struct, stores it in a per-database runtime record, and opens an
extension-owned DuckDB connection for native callback support. On extension
reload, Rducks verifies that the active connection has the SQL surface registered
and re-registers it when needed.

The unstable DuckDB C extension ABI entries are tracked by
`tools/used_duckdb_unstable_api.R` and documented in `docs/BUILD.md`. The README
intentionally does not source repository helper scripts while rendering. The
current output of
`source("tools/used_duckdb_unstable_api.R"); cat(rducks_used_duckdb_unstable_api_markdown("."))`
is:

| ABI group | Functions used | Count |
| --- | --- | ---: |
| `unstable_new_arrow_functions` | `duckdb_data_chunk_from_arrow`, `duckdb_data_chunk_to_arrow`, `duckdb_destroy_arrow_converted_schema`, `duckdb_schema_from_arrow`, `duckdb_to_arrow_schema` | 5 |
| `unstable_new_error_data_functions` | `duckdb_destroy_error_data`, `duckdb_error_data_has_error`, `duckdb_error_data_message` | 3 |
| `unstable_new_open_connect_functions` | `duckdb_client_context_get_connection_id`, `duckdb_connection_get_arrow_options`, `duckdb_destroy_arrow_options`, `duckdb_destroy_client_context` | 4 |
| `unstable_new_scalar_function_functions` | `duckdb_scalar_function_bind_get_extra_info`, `duckdb_scalar_function_bind_set_error`, `duckdb_scalar_function_get_client_context`, `duckdb_scalar_function_set_bind`, `duckdb_scalar_function_set_bind_data`, `duckdb_scalar_function_set_bind_data_copy` | 6 |
| `unstable_new_scalar_function_state_functions` | `duckdb_scalar_function_get_state`, `duckdb_scalar_function_init_get_bind_data`, `duckdb_scalar_function_init_get_client_context`, `duckdb_scalar_function_init_get_extra_info`, `duckdb_scalar_function_init_set_error`, `duckdb_scalar_function_init_set_state`, `duckdb_scalar_function_set_init` | 7 |
| `unstable_new_vector_functions` | `duckdb_create_selection_vector`, `duckdb_destroy_selection_vector`, `duckdb_selection_vector_get_data_ptr`, `duckdb_vector_copy_sel` | 4 |
