---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
options(width = 80)
library(Rducks)
```

# Rducks

[![R-CMD-check](https://github.com/sounkou-bioinfo/Rducks/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rducks/actions/workflows/R-CMD-check.yaml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Rducks)](https://sounkou-bioinfo.r-universe.dev/Rducks)

Rducks registers R functions as DuckDB SQL functions using a package-managed
DuckDB C extension. The extension records the DuckDB database instance at load
time and keeps extension-owned connections for native registration callbacks and
query streaming. It is built around explicit type descriptors, DuckDB Arrow C
Data/nanoarrow marshalling, and a strict rule that R object work runs on the
recorded R thread unless it is intentionally moved to R worker processes through
the Arrow IPC plan.

The user-facing surface separates DuckDB function kind, scalar-UDF evaluation
mode, execution plan, and R-side query consumption. Scalar UDFs use
`rducks_register_scalar_udf()` and choose `mode = "scalar"` for one R call per
row or `mode = "vectorized"` for one R call per DuckDB chunk. Execution plans
choose the marshalling/concurrency backend (`arrow_r`, `arrow_c`, or
`arrow_ipc`). Aggregates, table functions, and `rducks_query_stream()` are
separate APIs, not scalar-UDF plan variants.

## Quick start

```{r quick-start}
library(DBI)
library(duckdb)
library(Rducks)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(con, threads = "single")

score_udf <- rducks_register_scalar_udf(
  con,
  name = "r_score",
  fun = function(x, label) {
    bonus <- if (identical(label, "high")) 100 else 0
    as.double(x + bonus)
  },
  returns = DOUBLE,
  side_effects = TRUE
)

dbGetQuery(con, "
  SELECT r_score(x::DOUBLE, label) AS score
  FROM (VALUES (2, 'low'), (21, 'high')) AS t(x, label)
")
```

`r_score()` omits `args`, so DuckDB registers it as a dynamic varargs scalar
UDF. The return type is still explicit. At each SQL call site, DuckDB binds the
concrete argument types and Rducks materializes those inputs as if the signature
had been declared with `args = ...`. Use `args = NULL` only for a true
zero-argument UDF.

The returned registration object records the normalized signature and options;
DuckDB owns the catalog function after registration. Dropping the R object does
not unregister the SQL function. Registering the same SQL name/signature again
replaces the callable implementation. Use `side_effects = TRUE` for functions
with counters, randomness, I/O, mutation, sleeps, or other effects so DuckDB does
not optimize them as pure expressions.

## Lifecycle

`rducks_release(con)` detaches connection-local Rducks state and stops
Rducks-launched local IPC workers when the last attachment for the DuckDB runtime
is released. It does not drop DuckDB catalog functions or release closures still
owned by native catalog metadata. For deterministic cleanup, call it before
`DBI::dbDisconnect(con)`; to replace a scalar UDF, register the same SQL
name/signature again.

## Type descriptors

Rducks descriptors are used for scalar-UDF returns, declared scalar-UDF inputs,
and aggregate inputs/returns. They include DuckDB scalar types, exact value
classes such as `UUID`, `HUGEINT`, `DECIMAL(width, scale)`, `INTERVAL`, `BIT`,
`ENUM(levels)`, and composite descriptors such as `LIST(TYPE)`, `ARRAY(TYPE, n)`,
`STRUCT(...)`, `MAP(key, value)`, and `UNION(...)`.

```{r type-descriptors}
nested_type <- STRUCT(
  id = INTEGER,
  label = ENUM(c("low", "high")),
  payload = UNION(code = INTEGER, note = VARCHAR),
  values = LIST(DECIMAL(10, 2))
)

rducks_is_type(nested_type)
cat(strwrap(rducks_type_sql(nested_type), width = 70), sep = "\n")
rducks_type_child_names(nested_type)
```

## Scalar UDFs

Scalar mode calls the R function once per logical row. Vectorized mode calls the
R function once per DuckDB chunk with one R vector or list-column per declared or
dynamically bound argument.

```{r vectorized-scalar-udf}
scalar_plus_one_udf <- rducks_register_scalar_udf(
  con,
  name = "r_scalar_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "scalar",
  side_effects = TRUE
)

vec_plus_one_udf <- rducks_register_scalar_udf(
  con,
  name = "r_vec_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "vectorized",
  side_effects = TRUE
)

dbGetQuery(con, "SELECT sum(r_vec_plus_one(i::DOUBLE)) AS total FROM range(5) AS t(i)")
```

Dynamic omitted arguments are not a guessing path. They are bind-time descriptors.
The same R function below is registered once with an explicit nested signature
and once as dynamic varargs; both calls see the same typed R value.

```{r declared-vs-dynamic}
nested_summary <- function(x) {
  paste0(x$id, ":", x$label, ":", x$payload$tag, "=", x$payload$value)
}

nested_declared_udf <- rducks_register_scalar_udf(
  con,
  name = "r_nested_declared",
  fun = nested_summary,
  args = STRUCT(
    id = INTEGER,
    label = ENUM(c("low", "high")),
    payload = UNION(code = INTEGER, note = VARCHAR)
  ),
  returns = VARCHAR,
  null_handling = "special"
)

nested_dynamic_udf <- rducks_register_scalar_udf(
  con,
  name = "r_nested_dynamic",
  fun = nested_summary,
  returns = VARCHAR,
  null_handling = "special"
)

nested_sql <- "
  struct_pack(
    id := 7::INTEGER,
    label := 'high'::ENUM('low', 'high'),
    payload := union_value(note := 'ok')::UNION(code INTEGER, note VARCHAR)
  )
"

nested_query <- sprintf(
  paste(
    "SELECT",
    "  r_nested_declared(%1$s) AS declared,",
    "  r_nested_dynamic(%1$s) AS dynamic",
    sep = "\n"
  ),
  nested_sql
)
dbGetQuery(con, nested_query)
```

With `null_handling = "default"`, a top-level SQL `NULL` input produces a SQL
`NULL` output without calling R. With `null_handling = "special"`, top-level
SQL `NULL` values are passed as type-specific R missing values. Nested NULLs are
part of the value: scalar children usually become typed `NA`, while nested
composite NULLs become `NULL`.

```{r null-handling}
null_special_udf <- rducks_register_scalar_udf(
  con,
  name = "r_null_special",
  fun = function(x) if (is.na(x)) 5L else x,
  args = INTEGER,
  returns = INTEGER,
  null_handling = "special"
)

dbGetQuery(con, "SELECT r_null_special(NULL::INTEGER) AS x")
```

For type-by-type details, use the exported reference tables:

```{r semantic-tables}
rducks_mode_semantics()[, c("mode", "call_granularity", "input_shape")]
mapping <- rducks_argument_type_mapping(list(
  INTEGER,
  UUID,
  DECIMAL(10, 2),
  STRUCT(a = INTEGER[])
))
mapping[, c("duckdb_type", "r_value_class", "special_null_argument")]
```

## Aggregates, table functions, and query streams

### Aggregate functions

`rducks_register_aggregate()` registers R-backed DuckDB aggregates. Aggregate
state is an R object preserved by the extension, not a serialized blob. The
callbacks run on the recorded R thread and are not controlled by scalar-UDF
execution plans.

```{r aggregate}
sum_i32_aggregate <- rducks_register_aggregate(
  con,
  name = "r_sum_i32",
  update = function(state, x) {
    if (is.null(state)) state <- 0L
    as.integer(state + x)
  },
  finalize = function(state) if (is.null(state)) NA_integer_ else state,
  args = INTEGER,
  returns = INTEGER
)

dbGetQuery(
  con,
  paste(
    "SELECT r_sum_i32(i) AS total",
    "FROM (VALUES (1::INTEGER), (2::INTEGER), (NULL::INTEGER)) t(i)"
  )
)
```

### Table functions

`rducks_register_table()` infers the number of SQL arguments from the R function
formals and registers those inputs as DuckDB `ANY`. The result can be a data
frame, a named list of columns, or `rducks_table_stream()` for scan-time batches.
Finite results are imported once during bind; stream results import batches as
DuckDB scans.

```{r table-function}
rows_table <- rducks_register_table(
  con,
  name = "r_rows",
  fun = function(n) data.frame(i = seq_len(as.integer(n))),
  chunk_size = 2L
)

dbGetQuery(con, "SELECT * FROM r_rows(3) ORDER BY i")

stream_rows_table <- rducks_register_table(
  con,
  name = "r_stream_rows",
  fun = function(n) {
    next_i <- 1L
    limit <- as.integer(n)
    rducks_table_stream(
      prototype = data.frame(i = integer()),
      next_batch = function(batch_size) {
        if (next_i > limit) return(NULL)
        hi <- min(limit, next_i + as.integer(batch_size) - 1L)
        out <- data.frame(i = seq.int(next_i, hi))
        next_i <<- hi + 1L
        out
      }
    )
  },
  chunk_size = 2L
)

dbGetQuery(con, "SELECT sum(i) AS total FROM r_stream_rows(5)")
```

### Query streams

`rducks_query_stream()` is for R callers that want explicit DuckDB-native query
batches instead of an eager `DBI::dbGetQuery()` result. The stream fetches native
DuckDB chunks and imports them through Arrow C Data/nanoarrow. `next_batch()`
returns data frames by default; `format = "record_batch"` returns an owned
nanoarrow record batch so materialization can happen later.

```{r query-stream}
stream <- rducks_query_stream(
  con,
  "SELECT i::INTEGER AS i FROM range(1, 6) t(i)",
  batch_size = 2L
)
stream$next_batch()
stream$next_batch()
stream$close()

record_stream <- rducks_query_stream(
  con,
  "SELECT i::INTEGER AS i FROM range(1, 4) t(i)",
  batch_size = 2L,
  format = "record_batch"
)
record_batch <- record_stream$next_batch()
class(record_batch)
record_stream$close()
```

## Execution plans

Execution plans are fixed at scalar-UDF registration time.

| Plan | Scalar mode | Vectorized mode | Notes |
| --- | --- | --- | --- |
| `arrow_r + serial` | yes | yes | reference path using DuckDB Arrow C Data plus nanoarrow/R |
| `arrow_r + inproc_concurrent` | yes | yes | DuckDB workers enqueue callbacks; R work drains on the recorded R thread |
| `arrow_c + serial` | yes | yes | direct native DuckDB-vector materialization for supported types |
| `arrow_c + inproc_concurrent` | yes | yes | native materialization plus same-process queueing |
| `arrow_ipc + multiprocess_parallel` | yes | yes | native NNG plus owned Arrow IPC request/result bytes and R workers |

The benchmark below registers the same sleeping vectorized UDF on three real
plans and runs the queries against small CSV files. The UDF closes over a random
R lookup vector; the Arrow IPC registration sends that global explicitly using
`ipc_globals_share = "mori"`. Timings are illustrative and machine-dependent,
but the code exercises the actual `arrow_r`, `arrow_c`, and native NNG/Arrow IPC
paths.

```{r execution-plan-benchmark}
set.seed(1)
lookup <- sample.int(20L, 1000L, replace = TRUE)
slow_lookup <- function(x) {
  Sys.sleep(0.02)
  x + lookup[[1L]]
}

csv_dir <- file.path(tempdir(), paste0("rducks-readme-csv-", Sys.getpid()))
dir.create(csv_dir, showWarnings = FALSE)
for (part in seq_len(4L)) {
  values <- seq.int((part - 1L) * 256L, part * 256L - 1L)
  writeLines(
    c("i", as.character(values)),
    file.path(csv_dir, sprintf("part-%02d.csv", part))
  )
}
csv_glob <- file.path(csv_dir, "part-*.csv")

plans <- list(
  arrow_r = rducks_execution_plan("arrow_r", "serial"),
  arrow_c = rducks_execution_plan("arrow_c", "serial"),
  arrow_ipc_mori = rducks_execution_plan(
    "arrow_ipc", "multiprocess_parallel",
    ipc_workers = 2L,
    ipc_transport = "tcp",
    ipc_timeout = 30,
    ipc_globals = "lookup",
    ipc_globals_share = "mori"
  )
)
udfs <- paste0("r_bench_", names(plans))

for (i in seq_along(plans)) {
  rducks_set_execution_plan(con, plans[[i]], threads = 1, external_threads = 1)
  rducks_register_scalar_udf(
    con,
    name = udfs[[i]],
    fun = slow_lookup,
    args = INTEGER,
    returns = INTEGER,
    mode = "vectorized",
    side_effects = TRUE
  )
}

run_plan <- function(label, udf, plan, threads) {
  rducks_set_execution_plan(con, plan, threads = threads, external_threads = 1)
  elapsed <- system.time({
    result <- DBI::dbGetQuery(con, sprintf(
      "SELECT sum(%s(i::INTEGER)) AS total FROM read_csv_auto(%s, header = true)",
      DBI::dbQuoteIdentifier(con, udf),
      DBI::dbQuoteString(con, csv_glob)
    ))
  })[["elapsed"]]
  data.frame(label = label, total = result$total[[1]], elapsed_sec = round(elapsed, 3))
}

benchmark <- rbind(
  run_plan("arrow_r serial", udfs[[1]], plans[[1]], threads = 1),
  run_plan("arrow_c serial", udfs[[2]], plans[[2]], threads = 1),
  run_plan("arrow_ipc + mori", udfs[[3]], plans[[3]], threads = 2)
)
unlink(csv_dir, recursive = TRUE, force = TRUE)
rducks_set_execution_plan(
  con,
  rducks_execution_plan("arrow_r", "serial"),
  threads = 1,
  external_threads = 1
)
benchmark
```

## duckplyr integration

`rducks_with_duckplyr()` and the `with.duckdb_connection()` method let ordinary R
calls inside duckplyr expressions resolve to Rducks scalar UDFs, without writing
`dd$...` calls. The fallback-blocking demonstration lives in
`inst/examples/no_fallback_duckplyr.R`; the README keeps the shape minimal.

```{r duckplyr}
demo_con <- DBI::dbConnect(
  duckdb::duckdb(config = list(allow_unsigned_extensions = "true")),
  dbdir = ":memory:"
)
Rducks::rducks_enable(demo_con, threads = "single")
DBI::dbWriteTable(demo_con, "scores", data.frame(
  id = 1:3,
  x = c(2, 21, 34),
  label = c("low", "high", "high")
))

scores <- duckplyr::read_sql_duckdb(
  "SELECT * FROM scores",
  con = demo_con,
  prudence = "stingy"
)
local_score <- function(x, label) {
  as.double(x + if (identical(label, "high")) 100 else 0)
}

out <- with(
  demo_con,
  scores |>
    dplyr::mutate(score = local_score(x, label)) |>
    dplyr::select(id, score) |>
    dplyr::collect(),
  rducks_returns = list(local_score = DOUBLE)
)
out

Rducks::rducks_release(demo_con)
DBI::dbDisconnect(demo_con, shutdown = TRUE)
```

## Arrow conversion options

`rducks_enable()` sets `arrow_lossless_conversion=true` on the user connection,
and the extension sets the same option on its internal DuckDB connections. That
is required for Rducks' typed boundary: DuckDB-specific logical types such as
`UUID`, `HUGEINT`, `UHUGEINT`, `INTERVAL`, `BIT`, and enums must keep their type
metadata when chunks cross DuckDB Arrow C Data. Without that setting, Arrow
conversion can erase type identity and dynamic omitted-`args` calls would no
longer be equivalent to explicit descriptors.

## Build notes

The source and vendored native dependencies used by `configure` live under
`tools/ext/`. During source-package installation, `configure` writes the generated
artifact to `inst/rducks_extension/build/rducks.duckdb_extension` in the build
tree; after installation the runtime path is
`rducks_extension/build/rducks.duckdb_extension` inside the installed package.
`cleanup` removes only the generated artifact, not the source tree needed by
`R CMD build`. DuckDB C API headers are refreshed explicitly when the supported
DuckDB version changes.

```sh
Rscript tools/fetch_duckdb_headers.R --ref v1.5.2
```

The extension uses DuckDB's unstable C extension ABI for Arrow conversion,
connection/runtime access, scalar-function bind/init hooks, dynamic bind-time
argument inspection, and selection-vector helpers. This table is generated from
`tools/used_duckdb_unstable_api.R` when `README.Rmd` is rendered:

```{r unstable-api-table, echo=FALSE, results='asis'}
source("tools/used_duckdb_unstable_api.R")
cat(rducks_used_duckdb_unstable_api_markdown("."))
```

See `docs/BUILD.md` for the build and ABI details.

```{r cleanup, include=FALSE}
rducks_release(con)
dbDisconnect(con, shutdown = TRUE)
```
