---
title: "Rducks scalar UDF benchmark and profiling suite"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  sf: 0.01
  reps: 3
  warmup: 1
  micro_n: [10000, 100000, 300000]
  leak_iters: 30
  profile_rows: 100000
  run_micro: true
  run_tpch: true
  run_leak: true
  run_profile: true
  duckdb_json_profile: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Scope

This benchmark exercises the real Rducks public path:

- R package registration through `rducks_register()`.
- DuckDB scalar-function execution on real DuckDB chunks.
- `eval_mode = "R"` and `eval_mode = "RC"` with `side_effects = TRUE`.
- Direct-buffer RC paths for ordinary and exact/exotic scalar storage.
- Arrow/nanoarrow fallback paths for nested and composite types.
- TPC-H data generated by DuckDB's `tpch` extension when available.

It is deliberately **not** a package test. It is a profiling and regression suite
for local performance work. Defaults are small enough for development; increase
`params$sf`, `params$reps`, `params$micro_n`, and `params$leak_iters` for serious
runs.

Example:

```sh
Rscript -e 'rmarkdown::render(
  "tools/benchmark_rducks.Rmd",
  output_dir = "bench-results",
  params = list(sf = 0.1, reps = 7, micro_n = c(1e5, 1e6, 3e6), leak_iters = 100)
)'
```

For native leak checking, render a tiny run once, then run the same workload under
Valgrind or an ASAN-built R/DuckDB/Rducks stack. The in-process memory probes below
are trend detectors, not proof of leak freedom.

# Utilities

```{r utilities}
suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
  library(Rducks)
})

bench_started <- Sys.time()
bench_stamp <- format(bench_started, "%Y%m%d-%H%M%S")
bench_dir <- file.path("bench-results", paste0("rducks-", bench_stamp))
dir.create(bench_dir, recursive = TRUE, showWarnings = FALSE)

sql_string <- function(x) {
  paste0("'", gsub("'", "''", x, fixed = TRUE), "'")
}

rss_mb <- function() {
  path <- "/proc/self/status"
  if (!file.exists(path)) return(NA_real_)
  line <- grep("^VmRSS:", readLines(path, warn = FALSE), value = TRUE)
  if (!length(line)) return(NA_real_)
  as.numeric(sub("^VmRSS:[[:space:]]*([0-9]+)[[:space:]]*kB.*$", "\\1", line)) / 1024
}

gc_snapshot <- function() {
  g <- gc()
  c(
    ncells_used = unname(g["Ncells", "used"]),
    vcells_used = unname(g["Vcells", "used"])
  )
}

duckdb_memory_mb <- function(con) {
  tryCatch({
    x <- DBI::dbGetQuery(con, "SELECT sum(memory_usage_bytes) AS bytes FROM duckdb_memory()")
    as.numeric(x$bytes[[1]]) / 1024^2
  }, error = function(e) NA_real_)
}

memory_snapshot <- function(con) {
  g <- gc_snapshot()
  data.frame(
    rss_mb = rss_mb(),
    duckdb_memory_mb = duckdb_memory_mb(con),
    ncells_used = g[["ncells_used"]],
    vcells_used = g[["vcells_used"]],
    stringsAsFactors = FALSE
  )
}

connect_rducks <- function() {
  con <- DBI::dbConnect(duckdb::duckdb(config = list(allow_unsigned_extensions = "true")))
  # The benchmark intentionally exercises the supported calling-R-thread path.
  # Set these explicitly here as benchmark setup, not as hidden package behavior.
  DBI::dbExecute(con, "SET external_threads=1")
  DBI::dbExecute(con, "PRAGMA threads=1")
  rducks_enable(con, threads = "single")
  DBI::dbExecute(con, "SET external_threads=1")
  DBI::dbExecute(con, "PRAGMA threads=1")
  con
}

disconnect_rducks <- function(con) {
  try(DBI::dbDisconnect(con, shutdown = TRUE), silent = TRUE)
  invisible(NULL)
}

bench_sql <- function(con, label, sql, reps = params$reps, warmup = params$warmup) {
  for (i in seq_len(warmup)) invisible(DBI::dbGetQuery(con, sql))
  rows <- vector("list", reps)
  for (i in seq_len(reps)) {
    invisible(gc())
    before <- memory_snapshot(con)
    timing <- system.time(result <- DBI::dbGetQuery(con, sql))
    after_query <- memory_snapshot(con)
    invisible(gc())
    after_gc <- memory_snapshot(con)
    rows[[i]] <- data.frame(
      label = label,
      rep = i,
      elapsed_sec = unname(timing[["elapsed"]]),
      user_sec = unname(timing[["user.self"]]),
      system_sec = unname(timing[["sys.self"]]),
      result_rows = NROW(result),
      rss_before_mb = before$rss_mb,
      rss_after_query_mb = after_query$rss_mb,
      rss_after_gc_mb = after_gc$rss_mb,
      duckdb_memory_before_mb = before$duckdb_memory_mb,
      duckdb_memory_after_query_mb = after_query$duckdb_memory_mb,
      duckdb_memory_after_gc_mb = after_gc$duckdb_memory_mb,
      ncells_before = before$ncells_used,
      ncells_after_gc = after_gc$ncells_used,
      vcells_before = before$vcells_used,
      vcells_after_gc = after_gc$vcells_used,
      stringsAsFactors = FALSE
    )
  }
  do.call(rbind, rows)
}

summarise_bench <- function(x) {
  if (!NROW(x)) return(x)
  labels <- unique(x$label)
  do.call(rbind, lapply(labels, function(label) {
    y <- x[x$label == label, , drop = FALSE]
    data.frame(
      label = label,
      reps = NROW(y),
      min_sec = min(y$elapsed_sec),
      median_sec = median(y$elapsed_sec),
      mean_sec = mean(y$elapsed_sec),
      max_sec = max(y$elapsed_sec),
      rss_delta_after_gc_mb = tail(y$rss_after_gc_mb, 1) - y$rss_before_mb[[1]],
      duckdb_mem_delta_after_gc_mb = tail(y$duckdb_memory_after_gc_mb, 1) - y$duckdb_memory_before_mb[[1]],
      stringsAsFactors = FALSE
    )
  }))
}

write_bench <- function(x, name) {
  path <- file.path(bench_dir, paste0(name, ".csv"))
  utils::write.csv(x, path, row.names = FALSE)
  path
}

run_rprof <- function(con, label, sql, interval = 0.001) {
  prof_file <- file.path(bench_dir, paste0(label, ".Rprof"))
  utils::Rprof(prof_file, interval = interval, memory.profiling = TRUE)
  on.exit(utils::Rprof(NULL), add = TRUE)
  invisible(DBI::dbGetQuery(con, sql))
  utils::Rprof(NULL)
  summary <- tryCatch(
    utils::summaryRprof(prof_file, memory = "both"),
    error = function(e) utils::summaryRprof(prof_file)
  )
  by_self <- as.data.frame(summary$by.self)
  by_self$function_name <- rownames(by_self)
  rownames(by_self) <- NULL
  by_self <- by_self[, c("function_name", setdiff(names(by_self), "function_name")), drop = FALSE]
  utils::write.csv(by_self, file.path(bench_dir, paste0(label, "-rprof-by-self.csv")), row.names = FALSE)
  by_self
}

run_duckdb_json_profile <- function(con, label, sql) {
  path <- file.path(bench_dir, paste0(label, ".duckdb-profile.json"))
  DBI::dbExecute(con, paste0("PRAGMA profiling_output=", sql_string(normalizePath(path, mustWork = FALSE))))
  DBI::dbExecute(con, "PRAGMA enable_profiling='json'")
  on.exit(try(DBI::dbExecute(con, "PRAGMA disable_profiling"), silent = TRUE), add = TRUE)
  invisible(DBI::dbGetQuery(con, sql))
  DBI::dbExecute(con, "PRAGMA disable_profiling")
  path
}

explain_analyze <- function(con, label, sql) {
  out <- DBI::dbGetQuery(con, paste("EXPLAIN ANALYZE", sql))
  path <- file.path(bench_dir, paste0(label, "-explain-analyze.txt"))
  writeLines(capture.output(print(out, right = FALSE, row.names = FALSE)), path)
  path
}

params_used <- data.frame(
  parameter = names(params),
  value = vapply(params, function(x) paste(x, collapse = ","), character(1)),
  stringsAsFactors = FALSE
)
write_bench(params_used, "params")
params_used
```

# Buffer-reuse semantic guard

Before looking for faster buffer reuse, preserve the semantic boundary: Rducks may
reuse DuckDB-owned input and output buffers inside one DuckDB scalar-function
callback, but it must not reuse mutable R argument objects in a way that changes
objects retained by user code. This guard stores the first argument and first
`sys.call()` from a UDF invocation and checks that later rows do not mutate them.

```{r buffer-reuse-guard}
con_guard <- connect_rducks()

retained <- new.env(parent = emptyenv())
retained$arg <- NULL
retained$call <- NULL

invisible(rducks_register(
  con_guard,
  "rd_reuse_guard_rc",
  function(x) {
    if (is.null(retained$arg)) retained$arg <- x
    if (is.null(retained$call)) retained$call <- sys.call()
    x + 1L
  },
  INTEGER,
  INTEGER,
  eval_mode = "RC",
  side_effects = TRUE
))

guard_result <- DBI::dbGetQuery(con_guard, "SELECT rd_reuse_guard_rc(i::INTEGER) AS x FROM range(5) t(i)")
reuse_guard <- data.frame(
  output = paste(guard_result$x, collapse = ","),
  retained_arg = retained$arg,
  retained_arg_ok = identical(retained$arg, 0L),
  retained_call = paste(deparse(retained$call), collapse = " "),
  stringsAsFactors = FALSE
)
write_bench(reuse_guard, "buffer-reuse-guard")
reuse_guard
```

# Microbenchmarks

The microbenchmarks isolate scalar UDF overhead from table scan complexity. They
include native SQL baselines, direct-buffer RC cases for ordinary and
exact/exotic scalar storage, and fallback Arrow/nanoarrow cases for nested types.

```{r micro-register, eval=params$run_micro}
con_micro <- connect_rducks()

invisible(rducks_register(con_micro, "rd_bench_double_r", function(x) x + 1, DOUBLE, DOUBLE,
                          eval_mode = "R", side_effects = TRUE))
invisible(rducks_register(con_micro, "rd_bench_double_rc", function(x) x + 1, DOUBLE, DOUBLE,
                          eval_mode = "RC", side_effects = TRUE))
invisible(rducks_register(con_micro, "rd_bench_strlen_r", function(x) nchar(x), VARCHAR, INTEGER,
                          eval_mode = "R", side_effects = TRUE))
invisible(rducks_register(con_micro, "rd_bench_strlen_rc", function(x) nchar(x), VARCHAR, INTEGER,
                          eval_mode = "RC", side_effects = TRUE))
invisible(rducks_register(con_micro, "rd_bench_blob_r", function(x) c(x, as.raw(0xff)), BLOB, BLOB,
                          eval_mode = "R", side_effects = TRUE))
invisible(rducks_register(con_micro, "rd_bench_blob_rc", function(x) c(x, as.raw(0xff)), BLOB, BLOB,
                          eval_mode = "RC", side_effects = TRUE))
invisible(rducks_register(con_micro, "rd_bench_decimal_r", function(x) x + rducks_decimal("1.00", 15, 2), DECIMAL(15, 2), DECIMAL(15, 2),
                          eval_mode = "R", side_effects = TRUE))
invisible(rducks_register(con_micro, "rd_bench_decimal_rc", function(x) x + rducks_decimal("1.00", 15, 2), DECIMAL(15, 2), DECIMAL(15, 2),
                          eval_mode = "RC", side_effects = TRUE))
invisible(rducks_register(con_micro, "rd_bench_struct_r", function(x) x$a + x$b, STRUCT(a = INTEGER, b = INTEGER), INTEGER,
                          eval_mode = "R", side_effects = TRUE))
invisible(rducks_register(con_micro, "rd_bench_struct_rc", function(x) x$a + x$b, STRUCT(a = INTEGER, b = INTEGER), INTEGER,
                          eval_mode = "RC", side_effects = TRUE))
```

```{r micro-run, eval=params$run_micro}
micro_n <- as.integer(unlist(params$micro_n))
micro_rows <- list()
for (n in micro_n) {
  queries <- data.frame(
    label = c(
      sprintf("micro/native-double/n=%d", n),
      sprintf("micro/R-double-direct/n=%d", n),
      sprintf("micro/RC-double-direct/n=%d", n),
      sprintf("micro/R-varchar-direct/n=%d", n),
      sprintf("micro/RC-varchar-direct/n=%d", n),
      sprintf("micro/R-blob-direct/n=%d", n),
      sprintf("micro/RC-blob-direct/n=%d", n),
      sprintf("micro/R-decimal-arrow/n=%d", n),
      sprintf("micro/RC-decimal-direct/n=%d", n),
      sprintf("micro/R-struct-fallback/n=%d", n),
      sprintf("micro/RC-struct-fallback/n=%d", n)
    ),
    sql = c(
      sprintf("SELECT sum(i::DOUBLE + 1) AS x FROM range(%d) t(i)", n),
      sprintf("SELECT sum(rd_bench_double_r(i::DOUBLE)) AS x FROM range(%d) t(i)", n),
      sprintf("SELECT sum(rd_bench_double_rc(i::DOUBLE)) AS x FROM range(%d) t(i)", n),
      sprintf("SELECT sum(rd_bench_strlen_r(('duck' || (i %% 100)::VARCHAR)::VARCHAR)) AS x FROM range(%d) t(i)", n),
      sprintf("SELECT sum(rd_bench_strlen_rc(('duck' || (i %% 100)::VARCHAR)::VARCHAR)) AS x FROM range(%d) t(i)", n),
      sprintf("SELECT sum(octet_length(rd_bench_blob_r(from_hex('00AA')))) AS x FROM range(%d) t(i)", n),
      sprintf("SELECT sum(octet_length(rd_bench_blob_rc(from_hex('00AA')))) AS x FROM range(%d) t(i)", n),
      sprintf("SELECT sum(rd_bench_decimal_r((i %% 1000)::DECIMAL(15,2))::DOUBLE) AS x FROM range(%d) t(i)", n),
      sprintf("SELECT sum(rd_bench_decimal_rc((i %% 1000)::DECIMAL(15,2))::DOUBLE) AS x FROM range(%d) t(i)", n),
      sprintf("SELECT sum(rd_bench_struct_r({'a': i::INTEGER, 'b': 1::INTEGER}::STRUCT(a INTEGER, b INTEGER))) AS x FROM range(%d) t(i)", n),
      sprintf("SELECT sum(rd_bench_struct_rc({'a': i::INTEGER, 'b': 1::INTEGER}::STRUCT(a INTEGER, b INTEGER))) AS x FROM range(%d) t(i)", n)
    ),
    stringsAsFactors = FALSE
  )
  for (i in seq_len(NROW(queries))) {
    micro_rows[[length(micro_rows) + 1L]] <- bench_sql(con_micro, queries$label[[i]], queries$sql[[i]])
  }
}

micro_results <- do.call(rbind, micro_rows)
micro_summary <- summarise_bench(micro_results)
write_bench(micro_results, "micro-results")
write_bench(micro_summary, "micro-summary")
micro_summary
```

# TPC-H-inspired benchmark

This section uses DuckDB `dbgen` tables and TPC-H Q1-style aggregations. It is not
a full audited TPC-H submission; it is a realistic row-heavy workload for Rducks
scalar UDF execution.

```{r tpch-setup, eval=params$run_tpch}
con_tpch <- connect_rducks()

ensure_tpch <- function(con) {
  loaded <- tryCatch({ DBI::dbExecute(con, "LOAD tpch"); TRUE }, error = function(e) FALSE)
  if (!loaded) {
    loaded <- tryCatch({
      DBI::dbExecute(con, "INSTALL tpch")
      DBI::dbExecute(con, "LOAD tpch")
      TRUE
    }, error = function(e) {
      warning("DuckDB tpch extension unavailable: ", conditionMessage(e), call. = FALSE)
      FALSE
    })
  }
  loaded
}

tpch_available <- ensure_tpch(con_tpch)
if (tpch_available) {
  DBI::dbExecute(con_tpch, sprintf("CALL dbgen(sf=%s)", as.character(params$sf)))
  tpch_counts <- rbind(
    data.frame(table = "lineitem", rows = DBI::dbGetQuery(con_tpch, "SELECT count(*) AS n FROM lineitem")$n),
    data.frame(table = "orders", rows = DBI::dbGetQuery(con_tpch, "SELECT count(*) AS n FROM orders")$n),
    data.frame(table = "customer", rows = DBI::dbGetQuery(con_tpch, "SELECT count(*) AS n FROM customer")$n)
  )
  write_bench(tpch_counts, "tpch-counts")
  tpch_counts
}
```

```{r tpch-register, eval=params$run_tpch}
if (tpch_available) {
  invisible(rducks_register(con_tpch, "rd_tpch_disc_r", function(price, discount) price * (1 - discount), c(DOUBLE, DOUBLE), DOUBLE,
                            eval_mode = "R", side_effects = TRUE))
  invisible(rducks_register(con_tpch, "rd_tpch_disc_rc", function(price, discount) price * (1 - discount), c(DOUBLE, DOUBLE), DOUBLE,
                            eval_mode = "RC", side_effects = TRUE))
  invisible(rducks_register(con_tpch, "rd_tpch_charge_r", function(price, discount, tax) price * (1 - discount) * (1 + tax), c(DOUBLE, DOUBLE, DOUBLE), DOUBLE,
                            eval_mode = "R", side_effects = TRUE))
  invisible(rducks_register(con_tpch, "rd_tpch_charge_rc", function(price, discount, tax) price * (1 - discount) * (1 + tax), c(DOUBLE, DOUBLE, DOUBLE), DOUBLE,
                            eval_mode = "RC", side_effects = TRUE))
  invisible(rducks_register(con_tpch, "rd_tpch_ship_ok_r", function(shipdate) shipdate <= as.Date("1998-09-02"), DATE, BOOLEAN,
                            eval_mode = "R", side_effects = TRUE))
  invisible(rducks_register(con_tpch, "rd_tpch_ship_ok_rc", function(shipdate) shipdate <= as.Date("1998-09-02"), DATE, BOOLEAN,
                            eval_mode = "RC", side_effects = TRUE))
  invisible(rducks_register(con_tpch, "rd_tpch_mode_len_r", function(x) nchar(x), VARCHAR, INTEGER,
                            eval_mode = "R", side_effects = TRUE))
  invisible(rducks_register(con_tpch, "rd_tpch_mode_len_rc", function(x) nchar(x), VARCHAR, INTEGER,
                            eval_mode = "RC", side_effects = TRUE))
  invisible(rducks_register(con_tpch, "rd_tpch_decimal_add_r", function(x) x + rducks_decimal("1.00", 15, 2), DECIMAL(15, 2), DECIMAL(15, 2),
                            eval_mode = "R", side_effects = TRUE))
  invisible(rducks_register(con_tpch, "rd_tpch_decimal_add_rc", function(x) x + rducks_decimal("1.00", 15, 2), DECIMAL(15, 2), DECIMAL(15, 2),
                            eval_mode = "RC", side_effects = TRUE))
}
```

```{r tpch-queries, eval=params$run_tpch}
if (tpch_available) {
  q1_native <- "
    SELECT
      l_returnflag,
      l_linestatus,
      sum(l_quantity::DOUBLE) AS sum_qty,
      sum(l_extendedprice::DOUBLE) AS sum_base_price,
      sum(l_extendedprice::DOUBLE * (1 - l_discount::DOUBLE)) AS sum_disc_price,
      sum(l_extendedprice::DOUBLE * (1 - l_discount::DOUBLE) * (1 + l_tax::DOUBLE)) AS sum_charge,
      avg(l_quantity::DOUBLE) AS avg_qty,
      avg(l_extendedprice::DOUBLE) AS avg_price,
      avg(l_discount::DOUBLE) AS avg_disc,
      count(*) AS count_order
    FROM lineitem
    WHERE l_shipdate <= DATE '1998-09-02'
    GROUP BY l_returnflag, l_linestatus
    ORDER BY l_returnflag, l_linestatus
  "

  q1_udf_r <- "
    SELECT
      l_returnflag,
      l_linestatus,
      sum(l_quantity::DOUBLE) AS sum_qty,
      sum(l_extendedprice::DOUBLE) AS sum_base_price,
      sum(rd_tpch_disc_r(l_extendedprice::DOUBLE, l_discount::DOUBLE)) AS sum_disc_price,
      sum(rd_tpch_charge_r(l_extendedprice::DOUBLE, l_discount::DOUBLE, l_tax::DOUBLE)) AS sum_charge,
      avg(l_quantity::DOUBLE) AS avg_qty,
      avg(l_extendedprice::DOUBLE) AS avg_price,
      avg(l_discount::DOUBLE) AS avg_disc,
      count(*) AS count_order
    FROM lineitem
    WHERE l_shipdate <= DATE '1998-09-02'
    GROUP BY l_returnflag, l_linestatus
    ORDER BY l_returnflag, l_linestatus
  "

  q1_udf_rc <- sub("rd_tpch_disc_r", "rd_tpch_disc_rc", q1_udf_r, fixed = TRUE)
  q1_udf_rc <- sub("rd_tpch_charge_r", "rd_tpch_charge_rc", q1_udf_rc, fixed = TRUE)

  q1_udf_rc_filter <- sub("l_shipdate <= DATE '1998-09-02'", "rd_tpch_ship_ok_rc(l_shipdate)", q1_udf_rc, fixed = TRUE)

  q_shipmode_native <- "
    SELECT l_shipmode, sum(length(l_shipmode)) AS mode_len, count(*) AS n
    FROM lineitem
    GROUP BY l_shipmode
    ORDER BY l_shipmode
  "
  q_shipmode_r <- "
    SELECT l_shipmode, sum(rd_tpch_mode_len_r(l_shipmode)) AS mode_len, count(*) AS n
    FROM lineitem
    GROUP BY l_shipmode
    ORDER BY l_shipmode
  "
  q_shipmode_rc <- sub("rd_tpch_mode_len_r", "rd_tpch_mode_len_rc", q_shipmode_r, fixed = TRUE)

  q_decimal_r <- "SELECT sum(rd_tpch_decimal_add_r(l_extendedprice)::DOUBLE) AS x FROM lineitem"
  q_decimal_rc <- "SELECT sum(rd_tpch_decimal_add_rc(l_extendedprice)::DOUBLE) AS x FROM lineitem"

  tpch_queries <- data.frame(
    label = c(
      "tpch/q1-native",
      "tpch/q1-R-direct-compute",
      "tpch/q1-RC-direct-compute",
      "tpch/q1-RC-direct-compute-and-filter",
      "tpch/shipmode-native",
      "tpch/shipmode-R-direct",
      "tpch/shipmode-RC-direct",
      "tpch/decimal-R-arrow",
      "tpch/decimal-RC-direct"
    ),
    sql = c(
      q1_native,
      q1_udf_r,
      q1_udf_rc,
      q1_udf_rc_filter,
      q_shipmode_native,
      q_shipmode_r,
      q_shipmode_rc,
      q_decimal_r,
      q_decimal_rc
    ),
    stringsAsFactors = FALSE
  )
  write_bench(tpch_queries, "tpch-query-list")
  tpch_queries[, "label", drop = FALSE]
}
```

```{r tpch-run, eval=params$run_tpch}
if (tpch_available) {
  tpch_results <- do.call(rbind, lapply(seq_len(NROW(tpch_queries)), function(i) {
    bench_sql(con_tpch, tpch_queries$label[[i]], tpch_queries$sql[[i]])
  }))
  tpch_summary <- summarise_bench(tpch_results)
  write_bench(tpch_results, "tpch-results")
  write_bench(tpch_summary, "tpch-summary")
  tpch_summary
}
```

```{r tpch-explain, eval=params$run_tpch}
if (tpch_available) {
  explain_files <- vapply(seq_len(NROW(tpch_queries)), function(i) {
    explain_analyze(con_tpch, gsub("[^A-Za-z0-9_-]", "-", tpch_queries$label[[i]]), tpch_queries$sql[[i]])
  }, character(1))
  data.frame(label = tpch_queries$label, explain_file = explain_files, stringsAsFactors = FALSE)
}
```

```{r tpch-duckdb-json-profile, eval=params$run_tpch && params$duckdb_json_profile}
if (tpch_available) {
  profile_files <- vapply(seq_len(NROW(tpch_queries)), function(i) {
    run_duckdb_json_profile(con_tpch, gsub("[^A-Za-z0-9_-]", "-", tpch_queries$label[[i]]), tpch_queries$sql[[i]])
  }, character(1))
  data.frame(label = tpch_queries$label, duckdb_profile_file = profile_files, stringsAsFactors = FALSE)
}
```

# Memory trend probes

These loops repeat already-registered UDFs. They are intended to catch obvious
per-call leaks in the execution path. They do not treat one-time registration
state as a leak, and they do not prove allocator memory must return to the OS.
Inspect the slope and the raw CSV in `bench_dir`.

```{r leak-probes, eval=params$run_leak}
con_leak <- connect_rducks()

invisible(rducks_register(con_leak, "rd_leak_double_rc", function(x) x + 1, DOUBLE, DOUBLE,
                          eval_mode = "RC", side_effects = TRUE))
invisible(rducks_register(con_leak, "rd_leak_decimal_rc", function(x) x + rducks_decimal("1.00", 15, 2), DECIMAL(15, 2), DECIMAL(15, 2),
                          eval_mode = "RC", side_effects = TRUE))
invisible(rducks_register(con_leak, "rd_leak_struct_rc", function(x) x$a + x$b, STRUCT(a = INTEGER, b = INTEGER), INTEGER,
                          eval_mode = "RC", side_effects = TRUE))

leak_probe <- function(con, label, sql, iterations) {
  rows <- vector("list", iterations)
  invisible(DBI::dbGetQuery(con, sql))
  for (i in seq_len(iterations)) {
    invisible(gc())
    before <- memory_snapshot(con)
    timing <- system.time(invisible(DBI::dbGetQuery(con, sql)))
    invisible(gc())
    after <- memory_snapshot(con)
    rows[[i]] <- data.frame(
      label = label,
      iter = i,
      elapsed_sec = unname(timing[["elapsed"]]),
      rss_before_mb = before$rss_mb,
      rss_after_gc_mb = after$rss_mb,
      duckdb_memory_before_mb = before$duckdb_memory_mb,
      duckdb_memory_after_gc_mb = after$duckdb_memory_mb,
      ncells_before = before$ncells_used,
      ncells_after_gc = after$ncells_used,
      vcells_before = before$vcells_used,
      vcells_after_gc = after$vcells_used,
      stringsAsFactors = FALSE
    )
  }
  do.call(rbind, rows)
}

leak_n <- max(10000L, min(as.integer(unlist(params$micro_n)), na.rm = TRUE))
leak_queries <- data.frame(
  label = c("leak/RC-double-direct", "leak/RC-decimal-direct", "leak/RC-struct-fallback"),
  sql = c(
    sprintf("SELECT sum(rd_leak_double_rc(i::DOUBLE)) AS x FROM range(%d) t(i)", leak_n),
    sprintf("SELECT sum(rd_leak_decimal_rc((i %% 1000)::DECIMAL(15,2))::DOUBLE) AS x FROM range(%d) t(i)", leak_n),
    sprintf("SELECT sum(rd_leak_struct_rc({'a': i::INTEGER, 'b': 1::INTEGER}::STRUCT(a INTEGER, b INTEGER))) AS x FROM range(%d) t(i)", leak_n)
  ),
  stringsAsFactors = FALSE
)

leak_results <- do.call(rbind, lapply(seq_len(NROW(leak_queries)), function(i) {
  leak_probe(con_leak, leak_queries$label[[i]], leak_queries$sql[[i]], params$leak_iters)
}))

leak_slopes <- do.call(rbind, lapply(split(leak_results, leak_results$label), function(x) {
  data.frame(
    label = x$label[[1]],
    iterations = NROW(x),
    rss_slope_mb_per_iter = unname(coef(lm(rss_after_gc_mb ~ iter, data = x))[["iter"]]),
    duckdb_mem_slope_mb_per_iter = unname(coef(lm(duckdb_memory_after_gc_mb ~ iter, data = x))[["iter"]]),
    ncells_slope_per_iter = unname(coef(lm(ncells_after_gc ~ iter, data = x))[["iter"]]),
    vcells_slope_per_iter = unname(coef(lm(vcells_after_gc ~ iter, data = x))[["iter"]]),
    stringsAsFactors = FALSE
  )
}))

write_bench(leak_results, "leak-results")
write_bench(leak_slopes, "leak-slopes")
leak_slopes
```

# R profiling

`Rprof()` sees the R side of the boundary: argument object creation visible to R,
user function cost, validation, S7 dispatch, and Arrow/nanoarrow fallback helper
cost. It does not replace native profilers for C/DuckDB internals.

```{r rprof, eval=params$run_profile}
con_prof <- connect_rducks()

profile_n <- as.integer(params$profile_rows)
invisible(rducks_register(con_prof, "rd_prof_double_r", function(x) x + 1, DOUBLE, DOUBLE,
                          eval_mode = "R", side_effects = TRUE))
invisible(rducks_register(con_prof, "rd_prof_double_rc", function(x) x + 1, DOUBLE, DOUBLE,
                          eval_mode = "RC", side_effects = TRUE))
invisible(rducks_register(con_prof, "rd_prof_decimal_rc", function(x) x + rducks_decimal("1.00", 15, 2), DECIMAL(15, 2), DECIMAL(15, 2),
                          eval_mode = "RC", side_effects = TRUE))

prof_queries <- data.frame(
  label = c("rprof-R-double-direct", "rprof-RC-double-direct", "rprof-RC-decimal-direct"),
  sql = c(
    sprintf("SELECT sum(rd_prof_double_r(i::DOUBLE)) AS x FROM range(%d) t(i)", profile_n),
    sprintf("SELECT sum(rd_prof_double_rc(i::DOUBLE)) AS x FROM range(%d) t(i)", profile_n),
    sprintf("SELECT sum(rd_prof_decimal_rc((i %% 1000)::DECIMAL(15,2))::DOUBLE) AS x FROM range(%d) t(i)", profile_n)
  ),
  stringsAsFactors = FALSE
)

rprof_tables <- lapply(seq_len(NROW(prof_queries)), function(i) {
  top <- run_rprof(con_prof, prof_queries$label[[i]], prof_queries$sql[[i]])
  top$query <- prof_queries$label[[i]]
  top
})
rprof_heads <- lapply(rprof_tables, head, 25L)
rprof_names <- Reduce(union, lapply(rprof_heads, names))
rprof_heads <- lapply(rprof_heads, function(x) {
  missing <- setdiff(rprof_names, names(x))
  for (name in missing) x[[name]] <- NA
  x[, rprof_names, drop = FALSE]
})
rprof_top <- do.call(rbind, rprof_heads)
write_bench(rprof_top, "rprof-top")
rprof_top
```

# Native leak/profiler commands

For C-level leaks and CPU attribution, use the rendered workload as a guide and
run a smaller parameter set under native tooling. Examples:

```sh
# Valgrind leak smoke test. Keep params tiny; R + DuckDB produce allocator noise.
R -d "valgrind --leak-check=full --show-leak-kinds=definite,indirect --track-origins=yes --num-callers=40" \
  -e 'rmarkdown::render("tools/benchmark_rducks.Rmd", output_dir="bench-results", params=list(sf=0.01, reps=1, micro_n=c(1000), leak_iters=3, run_tpch=FALSE, run_profile=FALSE))'

# Linux perf, useful for DuckDB and Rducks native frames when symbols are present.
perf record --call-graph dwarf -- \
  Rscript -e 'rmarkdown::render("tools/benchmark_rducks.Rmd", output_dir="bench-results", params=list(sf=0.1, reps=3, micro_n=c(1e5), leak_iters=10, run_profile=FALSE))'
perf report
```

# Session info

```{r session-info}
session <- sessionInfo()
writeLines(capture.output(session), file.path(bench_dir, "session-info.txt"))
session
data.frame(
  benchmark_artifacts = normalizePath(bench_dir, mustWork = FALSE),
  stringsAsFactors = FALSE
)
```

```{r cleanup, include=FALSE}
if (exists("con_prof", inherits = FALSE)) disconnect_rducks(con_prof)
if (exists("con_leak", inherits = FALSE)) disconnect_rducks(con_leak)
if (exists("con_tpch", inherits = FALSE)) disconnect_rducks(con_tpch)
if (exists("con_micro", inherits = FALSE)) disconnect_rducks(con_micro)
if (exists("con_guard", inherits = FALSE)) disconnect_rducks(con_guard)
```
